---
title: "Introduction to BATON: Bayesian Optimization for Clinical Trial Calibration"
author: "BATON Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BATON}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

The **BATON** package provides a modular framework for constrained Bayesian optimization of Bayesian adaptive clinical trial designs. It implements:

- **Heteroskedastic Gaussian process surrogates** for modeling operating characteristics
- **Constraint-aware acquisition functions** (Expected Constrained Improvement)
- **Multi-fidelity simulation** for efficient budget allocation
- **Sensitivity and covariance analyses** for design understanding

# Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("naimurashid/BATON")
```

```{r}
library(BATON)
```

# Quick Start Example

## 1. Define Your Simulator

The simulator is the core of your calibration workflow. It takes design parameters and returns operating characteristics (power, type I error, expected sample size, etc.).

### Basic Simulator Structure

This simulator models a **two-stage adaptive trial with early stopping for futility**.
At an interim look, the trial stops early if the observed response rate is too low to be
promising. Otherwise, enrollment continues to `n_max` and H0 is rejected if the final
response rate exceeds `efficacy_threshold`.

Note that in practice, BATON is designed for *computationally expensive* simulators
(e.g., survival trials with Bayesian posterior probability stopping rules that take
seconds per evaluation). The toy simulator below is intentionally fast for vignette
purposes, but the optimization workflow is identical.

```{r}
# Fixed assumptions about the trial (not optimized)
p0 <- 0.20  # null response rate
p1 <- 0.40  # alternative response rate

# Two-stage adaptive trial simulator
simple_simulator <- function(theta, fidelity = c("low", "med", "high"), seed = NULL, ...) {
  fidelity <- match.arg(fidelity)

  # Number of Monte Carlo replications by fidelity
  n_rep <- switch(fidelity,
                  low = 200,
                  med = 1000,
                  high = 10000)

  if (!is.null(seed)) set.seed(seed)

  # Extract design parameters
  n_max              <- round(theta$n_max)
  interim_fraction   <- theta$interim_fraction
  futility_threshold <- theta$futility_threshold
  efficacy_threshold <- theta$efficacy_threshold

  n_interim <- max(1L, round(interim_fraction * n_max))
  n_final   <- n_max - n_interim

  # --- Vectorised simulation (no loop) ---

  sim_one_hypothesis <- function(p) {
    # Stage 1: interim responses
    x_interim  <- rbinom(n_rep, n_interim, p)
    not_futile <- (x_interim / n_interim) >= futility_threshold

    # Stage 2: remaining responses (only for non-futile trials)
    x_remaining <- rbinom(n_rep, n_final, p)
    x_total     <- x_interim + x_remaining

    reject <- not_futile & (x_total / n_max > efficacy_threshold)
    en     <- ifelse(not_futile, n_max, n_interim)

    list(reject = reject, en = en)
  }

  alt  <- sim_one_hypothesis(p1)
  null <- sim_one_hypothesis(p0)

  metrics <- c(
    power = mean(alt$reject),
    type1 = mean(null$reject),
    EN    = mean(alt$en)
  )

  variance <- c(
    power = var(alt$reject)  / n_rep,
    type1 = var(null$reject) / n_rep,
    EN    = var(alt$en)      / n_rep
  )

  attr(metrics, "variance") <- variance
  attr(metrics, "n_rep")    <- n_rep

  return(metrics)
}
```

### Test the Simulator

```{r}
# Test with some parameters
test_result <- simple_simulator(
  theta = list(n_max = 60, interim_fraction = 0.5,
               futility_threshold = 0.20, efficacy_threshold = 0.35),
  fidelity = "low",
  seed = 123
)

print(test_result)
print(attr(test_result, "variance"))
```

## 2. Define Design Space and Constraints

```{r}
# Parameter bounds
bounds <- list(
  n_max              = c(20, 100),   # Maximum sample size
  interim_fraction   = c(0.3, 0.7),  # Fraction enrolled at interim look
  futility_threshold = c(0.05, 0.40),# Stop early if interim rate below this
  efficacy_threshold = c(0.25, 0.60) # Reject H0 if final rate above this
)

# Objective: minimize expected sample size under the alternative
objective <- "EN"

# Constraints: power >= 0.80, type I error <= 0.10
constraints <- list(
  power = c("ge", 0.80),
  type1 = c("le", 0.10)
)
```

## 3. Run Bayesian Optimization

```{r eval=FALSE}
# Run calibration (d = 4; n_init = 10 x d, budget = ~15 x d)
fit <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  n_init = 40,      # Initial LHS design points (10 x d)
  q = 2,            # Batch size per iteration
  budget = 100,     # Total evaluations (~25 x d)
  seed = 2025
)
```

```{r echo=FALSE, eval=FALSE}
# Not evaluated at build time: running a live bo_calibrate() during
# build_vignettes = TRUE invokes a real Bayesian optimization, which is slow and
# can fail on a thin toolchain. The output is illustrative only.
fit <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  n_init = 20,
  q = 2,
  budget = 50,
  seed = 2025,
  progress = FALSE
)
```

## 4. Examine Results

```{r eval=FALSE}
# Not evaluated at build time: depends on `fit` from the (non-evaluated)
# calibration chunk above.
# Best design parameters
print(fit$best_theta)

# Operating characteristics at best design
feasible = fit$history[fit$history$feasible,]
best = feasible[which.min(feasible$objective),]
best_idx <- best$eval_id
print(fit$history$metrics[[best_idx]])

# Optimization history
head(fit$history[, c("iter", "fidelity", "objective", "feasible")])
```

# Memory-Efficient Variance Estimation

**Why variance matters**: Providing variance estimates enables heteroskedastic Gaussian process surrogates, which dramatically improve optimization efficiency (30-50% fewer evaluations).

## Using `welford_mean_var()` for Efficiency

Instead of storing all samples in memory, use Welford's online algorithm:

```{r}
# RECOMMENDED: Memory-efficient simulator using Welford's algorithm
efficient_simulator <- function(theta, fidelity = "high", seed = NULL, ...) {
  n_rep <- switch(fidelity,
                  low = 200,
                  med = 1000,
                  high = 10000)

  if (!is.null(seed)) set.seed(seed)

  # Use Welford's algorithm - no need to store all samples!
  result <- welford_mean_var(
    sample_fn = function(i, theta) {
      # Simulate ONE trial
      test_stat <- rnorm(1, mean = theta$threshold, sd = 1)
      power <- as.numeric(abs(test_stat) > qnorm(1 - theta$alpha/2))

      test_stat_null <- rnorm(1, mean = 0, sd = 1)
      type1 <- as.numeric(abs(test_stat_null) > qnorm(1 - theta$alpha/2))

      sample_size <- rpois(1, lambda = 100)

      c(power = power,
        type1 = type1,
        EN = sample_size,
        ET = sample_size / 10)
    },
    n_samples = n_rep,
    theta = theta
  )

  # Extract results
  metrics <- result$mean
  attr(metrics, "variance") <- result$variance
  attr(metrics, "n_rep") <- result$n

  return(metrics)
}
```

### Memory Comparison

```{r}
# Traditional approach (storing all samples)
traditional_memory <- function(n_rep, n_metrics) {
  8 * n_rep * n_metrics  # bytes
}

# Welford approach (incremental)
welford_memory <- function(n_metrics) {
  8 * 2 * n_metrics  # bytes (just mean and M2 vectors)
}

# For n_rep = 10,000 and 4 metrics:
cat("Traditional: ", traditional_memory(10000, 4) / 1024, "KB\n")
cat("Welford:     ", welford_memory(4), "bytes\n")
cat("Reduction:   ", traditional_memory(10000, 4) / welford_memory(4), "x\n")
```

**Result**: Welford uses **5,000x less memory** with only ~2% computational overhead!

## Parallel Simulators with `pool_welford_results()`

For expensive simulations, run chunks in parallel and pool the results:

```{r eval=FALSE}
parallel_simulator <- function(theta, fidelity = "high", n_cores = 4, ...) {
  n_rep <- switch(fidelity, low = 200, med = 1000, high = 10000)
  chunk_size <- ceiling(n_rep / n_cores)

  # Run chunks in parallel
  library(parallel)
  # NOTE: mclapply() is POSIX-only (no forking on Windows). Windows users should
  # use a PSOCK cluster (parallel::makeCluster() + parLapply()) or plain lapply().
  chunks <- mclapply(1:n_cores, function(core_id) {
    welford_mean_var(
      sample_fn = function(i, theta) {
        # Your trial simulation here
        c(power = ..., EN = ...)
      },
      n_samples = chunk_size,
      theta = theta
    )
  }, mc.cores = n_cores)

  # Pool results using Chan's algorithm
  pooled <- pool_welford_results(chunks)

  metrics <- pooled$mean
  attr(metrics, "variance") <- pooled$variance
  attr(metrics, "n_rep") <- pooled$n

  return(metrics)
}
```

# Multi-Fidelity Optimization

BATON automatically manages multi-fidelity evaluations to maximize efficiency:

## How It Works

1. **Initial design**: Low-fidelity evaluations explore the space
2. **Adaptive fidelity**: Higher fidelity for promising/uncertain regions
   - P(feasible) ≥ 0.75 → high fidelity (n=10,000)
   - P(feasible) ≥ 0.40 → medium fidelity (n=1,000)
   - Otherwise → low fidelity (n=200)
3. **Budget tracking**: Monitors total simulation cost

```{r}
# Examine fidelity allocation
table(fit$history$fidelity)

# Total simulation budget used
sum(fit$history$n_rep)
```

## Custom Fidelity Levels

```{r eval=FALSE}
fit_custom <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  # Custom fidelity levels
  fidelity_levels = c(
    low = 100,
    med = 500,
    high = 5000
  ),
  n_init = 10,
  budget = 50
)
```

# Benchmarking and Comparison

Compare Bayesian optimization against baseline strategies:

```{r eval=FALSE}
benchmark <- benchmark_methods(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  strategies = c("bo", "random", "grid"),
  bo_args = list(n_init = 10, q = 2, budget = 30, seeds = 1:3),
  random_args = list(n_samples = 30, seeds = 1:3),
  grid_args = list(resolution = list(threshold = 5, alpha = 5))
)

# Summary statistics
summary_tbl <- summarise_benchmark(benchmark)
print(summary_tbl)

# Visualization
plot_benchmark_trajectory(benchmark)
plot_benchmark_efficiency(benchmark)
```

# Sensitivity Analysis

Understand which parameters most influence your objective:

```{r eval=FALSE}
# Sobol sensitivity indices
sobol <- sa_sobol(
  surrogates = fit$surrogates,
  bounds = bounds,
  outcome = "EN",
  n_mc = 2000
)

print(sobol)
plot_sobol_indices(sobol)

# Local gradients
gradients <- sa_gradients(
  surrogates = fit$surrogates,
  theta = fit$best_theta,
  bounds = bounds,
  outcome = "EN"
)

print(gradients)

# Gradient covariance across design space
cov_matrix <- cov_effects(
  surrogates = fit$surrogates,
  bounds = bounds,
  outcome = "EN",
  n_mc = 500
)

print(cov_matrix)
```

# Advanced Topics

## Constraint Reliability

Validate that calibrated designs truly satisfy constraints:

```{r eval=FALSE}
reliability <- estimate_constraint_reliability(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  strategies = c("bo", "random"),
  calibration_seeds = 1:10,
  validation_reps = 50000,  # Large sample validation
  bo_args = list(n_init = 10, budget = 30)
)

print(reliability$summary)
plot_constraint_reliability(reliability)
```

## Multi-fidelity Ablation

Study the impact of different fidelity policies:

```{r eval=FALSE}
ablation <- ablation_multifidelity(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  policies = list(
    low_only = c(low = 200),
    med_only = c(med = 1000),
    full = c(low = 200, med = 1000, high = 10000)
  ),
  seeds = 1:10,
  bo_args = list(n_init = 10, budget = 30)
)

print(ablation$summary)
plot_multifidelity_tradeoff(ablation)
```

## Case Study Reporting

Extract publication-ready summaries:

```{r eval=FALSE}
# Summarize calibrated design
summary <- summarise_case_study(fit)
print(summary$design)
print(summary$operating_characteristics)

# Comprehensive diagnostics
diagnostics <- case_study_diagnostics(
  fit,
  sobol_samples = 5000,
  gradient_points = 200
)

# Visualizations
plot_feasible_frontier(fit)
plot_tradeoff_surfaces(fit)
plot_case_sobol(diagnostics$sobol)
plot_case_gradient(diagnostics$gradients)
plot_case_covariance(diagnostics$covariance)
```

# Best Practices

## ✅ DO:

1. **Always provide variance estimates** in your simulator
   - Use `welford_mean_var()` for memory efficiency
   - Enables heteroskedastic GP for 30-50% better performance

2. **Set appropriate fidelity levels**
   - Low: ~200 reps (exploration)
   - Med: ~1,000 reps (refinement)
   - High: ~10,000 reps (exploitation)

3. **Use sufficient initial design**
   - Minimum: `n_init ≥ 2 × d` where d = number of parameters
   - Recommended: `n_init = 4-5 × d`

4. **Monitor convergence**
   - Check `fit$history` for stagnation
   - Increase budget if needed

5. **Validate results**
   - Use `estimate_constraint_reliability()` with large samples
   - Confirm operating characteristics match expectations

## ⚠️ AVOID:

1. **Don't skip variance estimation**
   - Nugget fallback works but is 30-50% less efficient
   - For expensive simulations, this wastes significant time

2. **Don't use too small budgets**
   - Minimum: ~20-30 evaluations for 2D problems
   - Scale with dimensionality: ~10-15 × d

3. **Don't ignore infeasibility warnings**
   - If no feasible points found, relax constraints or expand bounds

4. **Don't over-trust surrogates far from data**
   - GP predictions are uncertain in unexplored regions
   - Use sensitivity analysis to understand uncertainty

# Troubleshooting

## Error: "Failed to fit surrogate for metric..."

**Cause**: GP fitting failed due to ill-conditioned covariance matrix

**Solutions**:
- Increase `n_init` for more diverse initial design
- Check for redundant/highly correlated evaluations
- Verify variance estimates are positive and finite

## Warning: "No feasible solutions found"

**Cause**: Constraints are too strict or bounds are wrong

**Solutions**:
- Check constraint thresholds are achievable
- Expand parameter bounds
- Increase budget to explore more

## Performance seems slow

**Check**:
1. Are you providing variance? (Check `attr(metrics, "variance")`)
2. Is your simulator the bottleneck? (Profile it)
3. Is `candidate_pool` too large? (Default 2000 is usually fine)

# Further Resources

- **Examples**: See `inst/examples/simulator_with_variance.R`
- **Code review fixes**: See `CODE_REVIEW_FIXES.md`
- **Variance explanation**: See `VARIANCE_ESTIMATOR_EXPLANATION.md`
- **Memory analysis**: See `MEMORY_ANALYSIS.md`
- **Performance impact**: See `NUGGET_IMPACT_ANALYSIS.md`
- **Development guide**: See `CLAUDE.md`

# Session Info

```{r}
sessionInfo()
```
