---
title: "Introduction to BATON: Bayesian Optimization for Clinical Trial Calibration"
author: "BATON Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BATON}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

The **BATON** package provides a modular framework for constrained Bayesian optimization of Bayesian adaptive clinical trial designs. It implements:

- **Heteroskedastic Gaussian process surrogates** for modeling operating characteristics
- **Constraint-aware acquisition functions** (Expected Constrained Improvement)
- **Multi-fidelity simulation** for efficient budget allocation
- **Sensitivity and covariance analyses** for design understanding

# Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("rashidlab/BATON")
```

```{r}
library(BATON)
```

# Quick Start Example

## 1. Define Your Simulator

The simulator is the core of your calibration workflow. It takes design parameters and returns operating characteristics (power, type I error, expected sample size, etc.).

### Basic Simulator Structure

This simulator models a **two-stage adaptive trial with early stopping for futility**.
At an interim look, the trial stops early if the observed response rate is too low to be
promising. Otherwise, enrollment continues to `n_max` and H0 is rejected if the final
response rate exceeds `efficacy_threshold`.

Note that in practice, BATON is designed for *computationally expensive* simulators
(e.g., survival trials with Bayesian posterior probability stopping rules that take
seconds per evaluation). The toy simulator below is intentionally fast for vignette
purposes, but the optimization workflow is identical.

```{r}
# Fixed assumptions about the trial (not optimized)
p0 <- 0.20  # null response rate
p1 <- 0.40  # alternative response rate

# Two-stage adaptive trial simulator
simple_simulator <- function(theta, fidelity = c("low", "med", "high"),
                             seed = NULL, n_rep = NULL, ...) {
  fidelity <- match.arg(fidelity)

  # BATON passes the exact requested replication count via n_rep. The switch()
  # is only a fallback for calls made outside BATON; keep it in sync with the
  # fidelity_levels used in bo_calibrate() (here the package defaults).
  if (is.null(n_rep)) {
    n_rep <- switch(fidelity,
                    low = 2000,
                    med = 4000,
                    high = 10000)
  }

  if (!is.null(seed)) set.seed(seed)

  # Extract design parameters
  n_max              <- round(theta$n_max)
  interim_fraction   <- theta$interim_fraction
  futility_threshold <- theta$futility_threshold
  efficacy_threshold <- theta$efficacy_threshold

  n_interim <- max(1L, round(interim_fraction * n_max))
  n_final   <- n_max - n_interim

  # --- Vectorised simulation (no loop) ---

  sim_one_hypothesis <- function(p) {
    # Stage 1: interim responses
    x_interim  <- rbinom(n_rep, n_interim, p)
    not_futile <- (x_interim / n_interim) >= futility_threshold

    # Stage 2: remaining responses (only for non-futile trials)
    x_remaining <- rbinom(n_rep, n_final, p)
    x_total     <- x_interim + x_remaining

    reject <- not_futile & (x_total / n_max > efficacy_threshold)
    en     <- ifelse(not_futile, n_max, n_interim)

    list(reject = reject, en = en)
  }

  alt  <- sim_one_hypothesis(p1)
  null <- sim_one_hypothesis(p0)

  metrics <- c(
    power = mean(alt$reject),
    type1 = mean(null$reject),
    EN    = mean(alt$en)
  )

  variance <- c(
    power = var(alt$reject)  / n_rep,
    type1 = var(null$reject) / n_rep,
    EN    = var(alt$en)      / n_rep
  )

  attr(metrics, "variance") <- variance
  attr(metrics, "n_rep")    <- n_rep

  return(metrics)
}
```

### Test the Simulator

```{r}
# Test with some parameters
test_result <- simple_simulator(
  theta = list(n_max = 60, interim_fraction = 0.5,
               futility_threshold = 0.20, efficacy_threshold = 0.35),
  fidelity = "low",
  seed = 123
)

print(test_result)
print(attr(test_result, "variance"))
```

## 2. Define Design Space and Constraints

```{r}
# Parameter bounds
bounds <- list(
  n_max              = c(20, 100),   # Maximum sample size
  interim_fraction   = c(0.3, 0.7),  # Fraction enrolled at interim look
  futility_threshold = c(0.05, 0.40),# Stop early if interim rate below this
  efficacy_threshold = c(0.25, 0.60) # Reject H0 if final rate above this
)

# Objective: minimize expected sample size under the alternative
objective <- "EN"

# Constraints: power >= 0.80, type I error <= 0.10
constraints <- list(
  power = c("ge", 0.80),
  type1 = c("le", 0.10)
)
```

Which hypothesis you minimize expected sample size under is a design decision:
minimizing `E[N]` under the null protects patients when the treatment does not
work, minimizing it under the alternative shortens trials of effective
treatments, and either is a legitimate objective. This vignette minimizes `EN`
under the alternative (the simulator returns `mean(alt$en)`); the README's
examples minimize it under the null.

## 3. Run Bayesian Optimization

The settings below follow the **Quick profile** from the README's "Run
Profiles" table (`n_init` = 4-5 x d, `budget` = 10-15 x d): appropriate for a
first exploratory run while you iterate on bounds, constraints, and metric
definitions. For a run whose answer you intend to act on, use the Balanced
profile (`n_init` = 10 x d, `budget` = 20-30 x d); for manuscript or
regulatory numbers, the Thorough profile. See the README for the full table.

```{r eval=FALSE}
# Run calibration (d = 4; Quick profile sizing)
fit <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  n_init = 20,      # Initial LHS design points (5 x d, Quick profile)
  q = 2,            # Batch size per iteration
  budget = 60,      # Total evaluations including n_init (15 x d, Quick profile)
  seed = 2025
)
```

## 4. Examine Results

```{r eval=FALSE}
# Not evaluated at build time: depends on `fit` from the (non-evaluated)
# calibration chunk above.
# Best design parameters and objective
print(fit$best_theta)
fit$best_objective

# Why the run ended: "budget_exhausted", "early_stopped", "acq_flatline", ...
fit$status

# Operating characteristics at the best design: the history tibble has one
# unpacked column per metric, so subset the best feasible row directly
feasible <- fit$history[fit$history$feasible, ]
best_row <- feasible[which.min(feasible$objective), ]
best_row[, c("power", "type1", "EN")]

# Optimization history
head(fit$history[, c("iter", "fidelity", "objective", "feasible")])
```

# Memory-Efficient Variance Estimation

**Why variance matters**: Providing variance estimates enables heteroskedastic Gaussian process surrogates, which dramatically improve optimization efficiency (30-50% fewer evaluations).

## Using `welford_mean_var()` for Efficiency

Instead of storing all samples in memory, use Welford's online algorithm:

```{r}
# RECOMMENDED: Memory-efficient simulator using Welford's algorithm
efficient_simulator <- function(theta, fidelity = "high", seed = NULL,
                                n_rep = NULL, ...) {
  # Fallback for calls outside BATON; keep in sync with your fidelity_levels
  if (is.null(n_rep)) {
    n_rep <- switch(fidelity,
                    low = 2000,
                    med = 4000,
                    high = 10000)
  }

  if (!is.null(seed)) set.seed(seed)

  # Use Welford's algorithm - no need to store all samples!
  result <- welford_mean_var(
    sample_fn = function(i, theta) {
      # Simulate ONE trial
      test_stat <- rnorm(1, mean = theta$threshold, sd = 1)
      power <- as.numeric(abs(test_stat) > qnorm(1 - theta$alpha/2))

      test_stat_null <- rnorm(1, mean = 0, sd = 1)
      type1 <- as.numeric(abs(test_stat_null) > qnorm(1 - theta$alpha/2))

      sample_size <- rpois(1, lambda = 100)

      c(power = power,
        type1 = type1,
        EN = sample_size,
        ET = sample_size / 10)
    },
    n_samples = n_rep,
    theta = theta
  )

  # Extract results
  metrics <- result$mean
  attr(metrics, "variance") <- result$variance
  attr(metrics, "n_rep") <- result$n

  return(metrics)
}
```

### Memory Comparison

```{r}
# Traditional approach (storing all samples)
traditional_memory <- function(n_rep, n_metrics) {
  8 * n_rep * n_metrics  # bytes
}

# Welford approach (incremental)
welford_memory <- function(n_metrics) {
  8 * 2 * n_metrics  # bytes (just mean and M2 vectors)
}

# For n_rep = 10,000 and 4 metrics:
cat("Traditional: ", traditional_memory(10000, 4) / 1024, "KB\n")
cat("Welford:     ", welford_memory(4), "bytes\n")
cat("Reduction:   ", traditional_memory(10000, 4) / welford_memory(4), "x\n")
```

**Result**: Welford uses **5,000x less memory**. The accumulator adds a small
per-sample cost that is negligible whenever simulating one trial does any real
work; see the variance-estimation vignette for a benchmark and an honest
reading of it.

## Parallel Simulators with `pool_welford_results()`

For expensive simulations, run chunks in parallel and pool the results:

```{r eval=FALSE}
parallel_simulator <- function(theta, fidelity = "high", n_cores = 4,
                               n_rep = NULL, ...) {
  # Fallback for calls outside BATON; keep in sync with your fidelity_levels
  if (is.null(n_rep)) {
    n_rep <- switch(fidelity, low = 2000, med = 4000, high = 10000)
  }
  chunk_size <- ceiling(n_rep / n_cores)

  # Run chunks in parallel
  library(parallel)
  # NOTE: mclapply() is POSIX-only (no forking on Windows). Windows users should
  # use a PSOCK cluster (parallel::makeCluster() + parLapply()) or plain lapply().
  chunks <- mclapply(1:n_cores, function(core_id) {
    welford_mean_var(
      sample_fn = function(i, theta) {
        # REPLACE with one replicate of YOUR trial; return a named vector,
        # e.g. c(power = ..., EN = ...)
        stop("REPLACE: simulate one replicate of your design")
      },
      n_samples = chunk_size,
      theta = theta
    )
  }, mc.cores = n_cores)

  # Pool results using Chan's algorithm
  pooled <- pool_welford_results(chunks)

  metrics <- pooled$mean
  attr(metrics, "variance") <- pooled$variance
  attr(metrics, "n_rep") <- pooled$n

  return(metrics)
}
```

# Multi-Fidelity Optimization

BATON automatically manages multi-fidelity evaluations to maximize efficiency:

## How It Works

By default BATON uses the `"adaptive"` fidelity method: for each proposed
design it computes a value score (acquisition value, objective uncertainty,
and proximity to the feasibility boundary) and divides by the tier's cost
raised to an exponent that grows as the run progresses and the budget
depletes. The tier with the best value per unit cost wins, so early
iterations spend cheap replications on exploration and late iterations
reserve expensive replications for verification.

The default replication counts are
`fidelity_levels = c(low = 2000, med = 4000, high = 10000)`; the README's
"Choosing Your Settings" table explains how to size them.

Four methods are available via `fidelity_method`:

| Method | One-line description |
|--------|----------------------|
| `"adaptive"` (default) | Cost-aware: maximizes expected information gain per unit cost |
| `"staged"` | Iteration-based schedule: low early, CV/boundary-triggered promotion mid-run, high late |
| `"threshold"` | Legacy: P(feasible) >= 0.75 high, >= 0.40 med, else low |
| `"hybrid_staged"` | MCEM-inspired staging with CV-based escalation and dynamic high-tier scaling |

```{r eval=FALSE}
# Examine fidelity allocation (depends on `fit` from the non-evaluated
# bo_calibrate chunk above)
table(fit$history$fidelity)

# Total simulation budget used
sum(fit$history$n_rep)
```

## Custom Fidelity Levels

```{r eval=FALSE}
fit_custom <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  # Custom fidelity levels. Because simple_simulator declares an n_rep
  # argument, BATON passes these exact counts to it (the simulator's internal
  # switch() fallback is bypassed).
  fidelity_levels = c(
    low = 100,
    med = 500,
    high = 5000
  ),
  n_init = 10,
  budget = 50
)
```

# Benchmarking and Comparison

Compare Bayesian optimization against baseline strategies:

```{r eval=FALSE}
benchmark <- benchmark_methods(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  strategies = c("bo", "random", "grid"),
  bo_args = list(n_init = 10, q = 2, budget = 30, seeds = 1:3),
  random_args = list(n_samples = 30, seeds = 1:3),
  # resolution names must match the names in `bounds`; misspelled names fall
  # back to the first listed resolution silently
  grid_args = list(resolution = list(n_max = 3, interim_fraction = 3,
                                     futility_threshold = 3,
                                     efficacy_threshold = 3))
)

# Summary statistics
summary_tbl <- summarise_benchmark(benchmark)
print(summary_tbl)

# Visualization
plot_benchmark_trajectory(benchmark)
plot_benchmark_efficiency(benchmark)
```

# Sensitivity Analysis

Understand which parameters most influence your objective:

```{r eval=FALSE}
# Sobol sensitivity indices
sobol <- sa_sobol(
  surrogates = fit$surrogates,
  bounds = bounds,
  outcome = "EN",
  n_mc = 2000
)

print(sobol)
plot_sobol_indices(sobol)

# Local gradients
gradients <- sa_gradients(
  surrogates = fit$surrogates,
  theta = fit$best_theta,
  bounds = bounds,
  outcome = "EN"
)

print(gradients)

# Gradient covariance across design space
cov_matrix <- cov_effects(
  surrogates = fit$surrogates,
  bounds = bounds,
  outcome = "EN",
  n_mc = 500
)

print(cov_matrix)
```

# Advanced Topics

## Constraint Reliability

Validate that calibrated designs truly satisfy constraints:

```{r eval=FALSE}
reliability <- estimate_constraint_reliability(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  strategies = c("bo", "random"),
  calibration_seeds = 1:10,
  validation_reps = 50000,  # Large sample validation
  bo_args = list(n_init = 10, budget = 30)
)

print(reliability$summary)
plot_constraint_reliability(reliability)
```

## Multi-fidelity Ablation

Study the impact of different fidelity policies:

```{r eval=FALSE}
ablation <- ablation_multifidelity(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  policies = list(
    low_only = c(low = 200),
    med_only = c(med = 1000),
    full = c(low = 200, med = 1000, high = 10000)
  ),
  seeds = 1:10,
  bo_args = list(n_init = 10, budget = 30)
)

print(ablation$summary)
plot_multifidelity_tradeoff(ablation)
```

## Case Study Reporting

Extract publication-ready summaries:

```{r eval=FALSE}
# Summarize calibrated design
summary <- summarise_case_study(fit)
print(summary$design)
print(summary$operating_characteristics)

# Comprehensive diagnostics
diagnostics <- case_study_diagnostics(
  fit,
  sobol_samples = 5000,
  gradient_points = 200
)

# Visualizations
plot_feasible_frontier(fit)
plot_tradeoff_surfaces(fit)
plot_case_sobol(diagnostics$sobol)
plot_case_gradient(diagnostics$gradients)
plot_case_covariance(diagnostics$covariance)
```

# Best Practices

## Do:

1. **Always provide variance estimates** in your simulator
   - Use `welford_mean_var()` for memory efficiency
   - Enables heteroskedastic GP for 30-50% better performance

2. **Set appropriate fidelity levels**
   - Defaults: low = 2,000, med = 4,000, high = 10,000 replications
   - Size the low tier cheap enough for exploration (it absorbs most
     evaluations) and the high tier precise enough for verification; see the
     README's "Choosing Your Settings" table

3. **Use sufficient initial design** (d = number of parameters)
   - Quick profile (first exploratory run): `n_init = 4-5 x d`
   - Balanced/Thorough (runs you act on): `n_init = 10 x d`
   - See the README's "Run Profiles" table for the full sizing guidance

4. **Monitor convergence**
   - Check `fit$history` for stagnation
   - Increase budget if needed

5. **Validate results**
   - Use `estimate_constraint_reliability()` with large samples
   - Confirm operating characteristics match expectations

## Avoid:

1. **Don't skip variance estimation**
   - Nugget fallback works but is 30-50% less efficient
   - For expensive simulations, this wastes significant time

2. **Don't use too small budgets**
   - Quick profile: `budget = 10-15 x d` for a first exploratory pass
   - Balanced: `20-30 x d`; Thorough: `40-50 x d` or more
   - Add 50-100% when constraints are tight (narrow feasible region)

3. **Don't ignore infeasibility warnings**
   - If no feasible points found, relax constraints or expand bounds

4. **Don't over-trust surrogates far from data**
   - GP predictions are uncertain in unexplored regions
   - Use sensitivity analysis to understand uncertainty

# Troubleshooting

## Error: "Failed to fit surrogate for metric..."

**Cause**: GP fitting failed due to ill-conditioned covariance matrix

**Solutions**:
- Increase `n_init` for more diverse initial design
- Check for redundant/highly correlated evaluations
- Verify variance estimates are positive and finite

## Warning: "No feasible solutions found"

**Cause**: Constraints are too strict or bounds are wrong

**Solutions**:
- Check constraint thresholds are achievable
- Expand parameter bounds
- Increase budget to explore more

## Performance seems slow

**Check**:
1. Are you providing variance? (Check `attr(metrics, "variance")`)
2. Is your simulator the bottleneck? (Profile it)
3. Is the candidate pool oversized? The pool is sized adaptively: 500 x d,
   clamped to [1000, 5000], with a 50% increase in the final 30% of
   iterations. The `candidate_pool` argument only sets a floor (default 2000).

# Before You Act on a Result

Before a calibrated design goes anywhere important, follow the README's
"Before You Trust the Answer" checklist: verify the best design at high
fidelity (`multi_seed_verify = TRUE`), round integer parameters and re-check
the operating characteristics, and inspect `fit$status` to see why the run
ended (`"budget_exhausted"`, `"early_stopped"`, `"acq_flatline"`, and the
service-control statuses). The README's "Run Profiles" table gives the
Quick/Balanced/Thorough settings referenced throughout this vignette.

# Further Resources

- **README**: "Choosing Your Settings", "Run Profiles", and "Before You
  Trust the Answer" sections (single source of truth for sizing guidance)
- **Other vignettes**: `vignette("advanced-features")`,
  `vignette("variance-estimation")`, `vignette("v04-hardening")`
- **Release notes**: `NEWS.md`
- **Examples**: `system.file("examples/simulator_with_variance.R", package = "BATON")`

# Session Info

```{r}
sessionInfo()
```
