---
title: "How BATON Works: Methods and Implementation"
author: "BATON Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How BATON Works: Methods and Implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
# Suppress the one-time q-default migration note so chunk output stays clean.
options(BATON.q_warning_shown = TRUE)
set.seed(2026)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r load}
library(BATON)
```

This vignette explains the statistical machinery inside `bo_calibrate()` and
how the implementation realizes it. It is written for readers who want to
understand what the package actually computes: which model is fit to the
simulation output, what the acquisition function rewards, how the
multi-fidelity budget is spent, and which implementation details (seeding,
deduplication, parallelism) matter for reproducibility. The presentation is
prose-first with a small number of key formulas, in the spirit of the
methods vignettes accompanying the `hetGP` and `DiceOptim` packages.

The methodology is developed in full in the companion manuscript (Young et
al., submitted; see the Citing section at the end). Everything below is
paraphrased from that manuscript and its supplement, checked line by line
against the package source, and where the two differ the description here
follows the code. Small runnable demonstrations are interleaved so the
abstractions stay concrete; each runs in a few seconds with fixed seeds.

For a task-oriented introduction (how to write a simulator, how to call
`bo_calibrate()`, how to read the results), start instead with
`vignette("BATON-introduction", package = "BATON")`.

# 1. The calibration problem

An adaptive trial design is parameterized by a vector
$\boldsymbol{\theta} \in \Theta \subset \mathbb{R}^d$ of calibration inputs:
sample size caps, decision thresholds, interim timing, prior
hyperparameters, and so on. For each design, the operating characteristics
(power, type I error, expected enrollment, expected duration) have no
closed form; they can only be estimated by simulating $R$ complete trials
and averaging. Write $\psi_m^*(\boldsymbol{\theta})$ for the true value of
the $m$-th operating characteristic and
$\hat{\psi}_m(\boldsymbol{\theta}; R)$ for its Monte Carlo estimate.

The calibration task BATON solves is a constrained minimization of one
operating characteristic subject to bounds on others:

$$
\min_{\boldsymbol{\theta} \in \Theta}\; \psi_0^*(\boldsymbol{\theta})
\qquad \text{subject to} \qquad
\psi_j^*(\boldsymbol{\theta}) \;\ge\; t_j
\;\;\text{or}\;\;
\psi_j^*(\boldsymbol{\theta}) \;\le\; t_j,
\qquad j = 1, \ldots, J,
$$

where $\Theta$ is a box (each parameter has a lower and upper bound, the
`bounds` argument), the objective $\psi_0$ is named by the `objective`
argument, and each constraint is a metric name with a direction (`"ge"` or
`"le"`) and a threshold (the `constraints` argument). A typical instance:
minimize expected enrollment subject to power $\ge 0.80$ and type I error
$\le 0.10$.

Four features distinguish this from textbook Bayesian optimization, and
each drives a design decision described in the sections that follow.

1. **The objective and the constraints are all noisy.** Every evaluation
   returns Monte Carlo estimates, $\hat{\psi}_m = \psi_m^* +
   \varepsilon_m$, with mean-zero error whose variance shrinks like $1/R$.
   A feasibility decision based on a noisy estimate can be wrong, so the
   search has to reason probabilistically about constraint satisfaction
   (Section 3).

2. **The noise is heteroskedastic and its variance is estimable.** For a
   proportion-valued metric estimated from $R$ replicates, the Monte Carlo
   variance is approximately $\hat{p}(1-\hat{p})/R$: near zero when the
   metric is near 0 or 1, maximal at 0.5. The simulator sees its own
   replicates, so it can hand BATON a per-evaluation variance estimate,
   and the surrogate model can use it (Section 2).

3. **Evaluations are expensive and batchable.** Simulating thousands of
   trials per design point dominates the cost of everything else, so the
   method budget is counted in simulator evaluations, points can be
   proposed in batches of `q` and evaluated in parallel, and the
   acquisition machinery must keep batch members from collapsing onto the
   same location (Section 3).

4. **Fidelity is a dial.** The same simulator run at $R = 200$ or $R =
   10{,}000$ replications answers the same question at different precision
   and cost. BATON treats the replication count as a decision variable
   (Section 4).

# 2. Gaussian process surrogates under Monte Carlo noise

## One independent GP per metric

`fit_surrogates()` fits one Gaussian process per modeled metric: the
objective plus every metric that appears in a constraint. The surrogates
are statistically independent of one another; no cross-metric covariance
is modeled. Joint feasibility probabilities are therefore products of
per-constraint marginals (Section 3), an approximation the manuscript
supports empirically for its application (constraint estimates come from
separate simulation scenarios, and observed posterior rank correlations
were small).

Each GP is fit on the unit hypercube: all parameters are rescaled to
$[0,1]$ internally (Section 6), so lengthscales are comparable across
dimensions and kernel defaults behave predictably.

## Heteroskedastic versus homoskedastic fits, and how BATON chooses

The Monte Carlo noise variance
$\mathrm{Var}(\hat{\psi}_m(\boldsymbol{\theta}; R)) =
\tau_m^2(\boldsymbol{\theta})/R$ depends on the input (through
$\tau_m^2$) and on the fidelity (through $R$). A homoskedastic surrogate
that assumes one common noise level will be overconfident exactly where
precision matters most, near constraint boundaries where proportion
metrics sit in their high-variance mid-range. BATON therefore prefers
noise-aware fits, with a graceful degradation chain. The selection logic
in `fit_surrogates()` is, per metric:

1. **hetGP path.** If the history contains *replicated designs* (the same
   $\boldsymbol{\theta}$ evaluated more than once, detected via the
   `theta_id` hash) *and* variance information is present, the metric is
   fit with `hetGP::mleHetGP()`. The replicate structure is passed
   directly (each observation is one row, with duplicated design
   locations), and hetGP estimates an input-dependent noise surface
   jointly with the mean surface, following Binois, Gramacy, and Ludkovski
   (2018). A small fixed nugget ($10^{-8}$) is added for numerical
   conditioning.
2. **DiceKriging path with known noise.** Without replicated designs (the
   common case mid-run, since duplicate proposals are rare in a continuous
   space), the metric is fit with `DiceKriging::km()` on the aggregated
   observations, passing the simulator-supplied Monte Carlo variances as
   the known per-observation noise (`noise.var`). This is still a
   heteroskedastic model in the sense that matters: each observation
   carries its own noise variance, so a high-fidelity evaluation is
   trusted more than a low-fidelity one at the fitting stage. Missing
   variances within a metric are imputed with the *largest* observed
   noise for that metric (an unknown-variance observation should be
   trusted least, not most).
3. **Homoskedastic nugget fallback.** If no variance information exists at
   all for a metric, `km()` is fit with a small fixed nugget
   ($10^{-6}$) instead, and a one-time message suggests attaching
   variance attributes to the simulator output.
4. **Constant predictor.** If every fitting attempt fails (which the
   hetGP path additionally guards with an `mleHomGP` retry and a `km`
   retry), the metric falls back to a constant predictor: the sample mean
   with a deliberately large predictive standard deviation (1.0), so the
   acquisition function treats the metric as unknown and keeps exploring
   rather than crashing the run.

The practical takeaway for simulator authors: attach
`attr(metrics, "variance")` (and `attr(metrics, "n_rep")`) to your
simulator's return value. If you do not, BATON falls back to a binomial
variance approximation $\hat{p}(1-\hat{p})/R$ for metrics that lie in
$[0,1]$ and to the homoskedastic nugget for everything else
(`default_variance_estimator()`), which is workable but weaker than a
proper replicate-based estimate.

## Aggregation of replicated designs

Evaluations sharing a `theta_id` (the same design up to a $10^{-6}$
rounding in unit coordinates, Section 6) are grouped before fitting. On
the DiceKriging path the grouped observations are averaged per metric
with NA-aware means, and their noise variances are averaged the same way,
so re-evaluating a promising design at higher fidelity sharpens the
surrogate there instead of creating a numerically degenerate duplicate
row. On the hetGP path replicates are passed through unaggregated, which
is precisely the structure hetGP exploits to separate signal variance
from noise variance.

## Estimating the variances: Welford and pooling

The per-evaluation variance estimates are the simulator author's
responsibility, and the package ships helpers that make this cheap:
`welford_mean_var()` computes means and variances in one streaming pass
over replicates (constant memory), and `pool_welford_results()` combines
chunked or parallel runs exactly. See
`vignette("variance-estimation", package = "BATON")` for the details and
worked examples.

## Kernel choice

The default covariance kernel is Matern 5/2 (`covtype = "matern5_2"`,
mapped to `Matern5_2` on the hetGP path), with automatic relevance
determination: one lengthscale per input dimension. Matern 5/2 sample
paths are twice differentiable, which matches the moderate smoothness of
simulated operating-characteristic surfaces better than the infinitely
smooth squared-exponential kernel, while remaining smooth enough for
stable maximum-likelihood fitting. This is the DiceKriging default and is
set explicitly for hetGP (whose own default is squared-exponential).

## Warm-starting fits across iterations

Surrogates are refit from scratch every BO iteration as the history
grows. To cut the cost of hyperparameter optimization,
`fit_surrogates()` accepts the previous iteration's fitted models and
extracts their lengthscales (`extract_gp_hyperparams()`: the
`range.val` slot of a `km` object, the `$theta` element of a hetGP or
homGP object). On the DiceKriging path these are passed as `parinit`,
the starting point for `km()`'s likelihood optimization, whenever they
form a valid positive vector of the right dimension; otherwise the fit
cold-starts. The hetGP path does not currently consume the previous
hyperparameters; its fits use hetGP's own initialization. Warm-starting
changes only the optimizer's starting point, not the model class, so it
trades a modest reduction in fitting time for no change in what is being
estimated. If a warm-started fit errors, `bo_calibrate()` retries the
iteration's fits without warm-start before giving up.

## Demonstration: what a surrogate is

The following builds a small history by hand (twelve noisy evaluations of
a one-dimensional function, each an average of 200 replicates with known
Monte Carlo variance), fits a surrogate through the exported
`fit_surrogates()`, and plots the GP mean and a two-standard-deviation
band against the truth. This is exactly the model `bo_calibrate()` fits
per metric; here the variance information routes the fit to the
DiceKriging known-noise path because no design is replicated.

```{r demo-gp-fit}
set.seed(101)
f_true <- function(x) sin(6 * x) + 0.6 * x

n_pts   <- 12
x_obs   <- seq(0.03, 0.97, length.out = n_pts)
R_reps  <- 200
sd_rep  <- 0.5   # per-replicate noise SD
y_obs   <- vapply(x_obs,
                  function(xi) mean(rnorm(R_reps, f_true(xi), sd_rep)),
                  numeric(1))
v_obs   <- rep(sd_rep^2 / R_reps, n_pts)   # known Monte Carlo variance

demo_history <- tibble::tibble(
  unit_x   = lapply(x_obs, function(xi) c(x = xi)),
  theta_id = vapply(x_obs, function(xi) BATON:::theta_to_id(c(x = xi)),
                    character(1)),
  metrics  = lapply(y_obs, function(yi) c(f = yi)),
  variance = lapply(v_obs, function(vi) c(f = vi))
)

no_constraints <- BATON:::parse_constraints(list())
sur <- fit_surrogates(demo_history, objective = "f",
                      constraint_tbl = no_constraints)
class(sur$f)   # DiceKriging km fit with per-point noise variances
```

```{r demo-gp-predict}
x_grid <- matrix(seq(0, 1, length.out = 201), ncol = 1,
                 dimnames = list(NULL, "x"))
pred <- BATON:::predict_surrogates(sur, x_grid)
gp_mean <- pred$f$mean
gp_sd   <- pred$f$sd
```

```{r demo-gp-plot, eval = has_ggplot2, fig.alt = "GP posterior mean with a two standard deviation band, the true function, and the noisy observations"}
library(ggplot2)
df_grid <- data.frame(x = x_grid[, "x"], mean = gp_mean, sd = gp_sd,
                      truth = f_true(x_grid[, "x"]))
df_obs  <- data.frame(x = x_obs, y = y_obs)
ggplot(df_grid, aes(x = x)) +
  geom_ribbon(aes(ymin = mean - 2 * sd, ymax = mean + 2 * sd),
              fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, linetype = "GP posterior mean"),
            color = "steelblue", linewidth = 0.9) +
  geom_line(aes(y = truth, linetype = "true function"), color = "black") +
  geom_point(data = df_obs, aes(y = y), size = 2) +
  scale_linetype_manual(NULL, values = c("GP posterior mean" = "solid",
                                         "true function" = "dashed")) +
  labs(x = "design parameter (unit scale)", y = "metric value",
       title = "A surrogate: GP mean and 2-sd band over noisy evaluations") +
  theme_minimal()
```

The band is the model's honest report of what it does not know: wide
between observations, and never collapsing to zero at the observations
themselves because each carries known Monte Carlo noise. Everything the
acquisition function does in the next section is a computation on this
mean and this band.

# 3. Acquisition: expected constrained improvement

## Expected improvement

Given the objective surrogate's posterior mean $\mu_0(\boldsymbol{\theta})$
and standard deviation $\sigma_0(\boldsymbol{\theta})$, and an incumbent
value $f^*$ (the best feasible objective known so far; BATON minimizes),
expected improvement (Jones, Schonlau, and Welch 1998) is

$$
\mathrm{EI}(\boldsymbol{\theta})
= \mathbb{E}\!\left[\max\{0,\; f^* - f(\boldsymbol{\theta})\}\right]
= \left(f^* - \mu_0\right)\Phi(z) + \sigma_0\,\phi(z),
\qquad
z = \frac{f^* - \mu_0}{\sigma_0},
$$

with $\Phi$ and $\phi$ the standard normal CDF and density. It is large
where the mean is already better than the incumbent, and it stays
positive where the mean is worse but the uncertainty leaves room for a
pleasant surprise. The implementation (`compute_ei()`, internal) adds an
epsilon guard of $10^{-10}$ to the standard deviation and handles the
degenerate $\sigma_0 \approx 0$ case as $\max\{0, f^* - \mu_0\}$.

## Feasibility probability

Because each constraint has its own independent GP, the probability that
constraint $j$ holds at $\boldsymbol{\theta}$ is a univariate Gaussian
tail probability,

$$
\pi_j(\boldsymbol{\theta}) =
\Phi\!\left(\frac{\mu_j(\boldsymbol{\theta}) - t_j}
{\sigma_j(\boldsymbol{\theta})}\right)
\;\;\text{for a "ge" constraint},
\qquad
\Phi\!\left(\frac{t_j - \mu_j(\boldsymbol{\theta})}
{\sigma_j(\boldsymbol{\theta})}\right)
\;\;\text{for "le"},
$$

and the joint feasibility probability is the product
$\Pi(\boldsymbol{\theta}) = \prod_{j=1}^J \pi_j(\boldsymbol{\theta})$
(`prob_feasibility_batch()`, vectorized across the whole candidate
pool). The product form follows Gelbart, Snoek, and Adams (2014) and
Gardner et al. (2014) and is exact when the constraint surfaces are
independent, an assumption discussed in Section 2.

## ECI

Expected constrained improvement multiplies the two:

$$
\alpha_{\mathrm{ECI}}(\boldsymbol{\theta})
= \mathrm{EI}(\boldsymbol{\theta}) \times \Pi(\boldsymbol{\theta}).
$$

A candidate scores highly only if it plausibly improves the objective
*and* plausibly satisfies every constraint. Since the optimum of a
constrained problem typically sits on the feasibility boundary, ECI
naturally concentrates evaluations there: inside the feasible region
$\Pi \approx 1$ and EI dominates; far outside, $\Pi \approx 0$ kills the
score regardless of how good the objective looks. The exported
implementation is `acq_eci()`.

## Before any feasible point exists

Early in a run with tight constraints, no evaluated design may be
feasible, so there is no incumbent and EI is undefined. BATON then
switches to an expected-violation criterion that steers the search toward
the feasible region. For each constraint, let $z_j$ be the standardized
*signed distance into the infeasible region* (for a "ge" constraint,
$z_j = (t_j - \mu_j)/\sigma_j$; for "le", $z_j = (\mu_j - t_j)/\sigma_j$).
The expected violation magnitude under the Gaussian posterior is

$$
\mathbb{E}\!\left[\max\{0, \text{violation}\}\right]
= \sigma_j \left( z_j\,\Phi(z_j) + \phi(z_j) \right),
$$

which grows without bound as a point becomes more deeply infeasible and
decays to zero well inside the feasible region, exactly the shape needed
to pull the search across the boundary. `compute_expected_violation()`
sums this over the constraints, and the pre-feasible acquisition is then

$$
\alpha_{\text{pre}}(\boldsymbol{\theta}) =
\left(1 - \tilde{v}(\boldsymbol{\theta}) + 0.3\,\sigma_0(\boldsymbol{\theta})\right)
\left(0.3 + 0.7\,\Pi(\boldsymbol{\theta})\right),
$$

where $\tilde{v}$ is the summed expected violation normalized by its
maximum over the candidate pool. The three ingredients read directly:
prefer small expected violation, keep a mild exploration bonus on
objective uncertainty, and weight by feasibility probability without
letting it hit zero. Once the first feasible design is observed, the
acquisition switches to standard ECI.

## Demonstration: ECI trading off improvement against feasibility

A one-dimensional constrained problem makes the trade visible: minimize
an expected-enrollment-like objective that increases in $x$, subject to a
power-like constraint that also increases in $x$. Cheap designs are
infeasible; feasible designs are expensive; the interesting region is the
boundary. The chunk fits both surrogates through `fit_surrogates()`, then
computes EI, $\Pi$, and ECI on a grid with the formulas above, inline for
pedagogy, and checks the result against BATON's `acq_eci()`.

```{r demo-eci-setup}
set.seed(202)
power_true <- function(x) plogis(9 * (x - 0.45))   # increasing in x
en_true    <- function(x) 30 + 45 * x              # increasing in x

n_pts  <- 14
R_reps <- 300
x_obs  <- seq(0.02, 0.98, length.out = n_pts)
pow_hat <- rbinom(n_pts, R_reps, power_true(x_obs)) / R_reps
en_hat  <- en_true(x_obs) + rnorm(n_pts, 0, 1.5)

eci_history <- tibble::tibble(
  unit_x   = lapply(x_obs, function(xi) c(x = xi)),
  theta_id = vapply(x_obs, function(xi) BATON:::theta_to_id(c(x = xi)),
                    character(1)),
  metrics  = lapply(seq_len(n_pts),
                    function(i) c(EN = en_hat[i], power = pow_hat[i])),
  variance = lapply(seq_len(n_pts),
                    function(i) c(EN = 1.5^2,
                                  power = max(pow_hat[i] * (1 - pow_hat[i]),
                                              0.01) / R_reps))
)

ctbl <- BATON:::parse_constraints(list(power = c("ge", 0.8)))
sur2 <- fit_surrogates(eci_history, objective = "EN", constraint_tbl = ctbl)

# Incumbent: best observed objective among observed-feasible designs
feasible_obs <- pow_hat >= 0.8
best_feas <- min(en_hat[feasible_obs])
round(best_feas, 1)
```

```{r demo-eci-compute}
pred2 <- BATON:::predict_surrogates(sur2, x_grid)
mu_EN <- pred2$EN$mean;    sd_EN <- pred2$EN$sd
mu_p  <- pred2$power$mean; sd_p  <- pred2$power$sd

# Inline formulas (BATON's implementation lives in acq_eci / compute_ei)
z_imp  <- (best_feas - mu_EN) / sd_EN
ei     <- (best_feas - mu_EN) * pnorm(z_imp) + sd_EN * dnorm(z_imp)
p_feas <- pnorm((mu_p - 0.8) / sd_p)
eci    <- ei * p_feas

# Same quantity through the package
eci_pkg <- acq_eci(x_grid, sur2, ctbl, objective = "EN",
                   best_feasible = best_feas)
all.equal(eci, eci_pkg, tolerance = 1e-6)
```

```{r demo-eci-plot, eval = has_ggplot2, fig.height = 5.5, fig.alt = "Three stacked panels showing expected improvement, feasibility probability, and their product ECI over the design parameter"}
df_eci <- rbind(
  data.frame(x = x_grid[, "x"], value = ei,     panel = "1. EI (objective)"),
  data.frame(x = x_grid[, "x"], value = p_feas, panel = "2. P(feasible)"),
  data.frame(x = x_grid[, "x"], value = eci,    panel = "3. ECI = EI x P(feasible)")
)
ggplot(df_eci, aes(x = x, y = value)) +
  geom_line(color = "steelblue", linewidth = 0.9) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(x = "design parameter (unit scale)", y = NULL,
       title = "ECI balances improvement against feasibility") +
  theme_minimal()
```

EI alone would send the next evaluation to small $x$, where the objective
is cheap but the power constraint is hopeless. The feasibility factor
suppresses that region, and the ECI maximum lands near the constraint
boundary, where a design might be both cheap and adequately powered. That
is the location worth spending simulations on.

## Batch selection by local penalization

When `q > 1`, BATON must propose several points from the *same* surrogate
state, and naively taking the top `q` acquisition scores returns `q`
near-copies of the same maximizer. BATON uses the local penalization
scheme of Gonzalez et al. (2016): select the acquisition maximizer, then
suppress a ball around it and re-maximize, repeating until the batch is
full. Concretely (`select_batch_local_penalization()`), after selecting
point $\boldsymbol{\theta}_k$ with acquisition value $\alpha_k$, every
remaining candidate at distance $r$ has its score reduced by

$$
\max\{0,\; \alpha_k - L\,r\},
$$

so the penalty is largest at the selected point and decays linearly to
zero at radius $\alpha_k / L$. The constant $L$ is a Lipschitz-style
smoothness estimate extracted from the objective GP
(`estimate_lipschitz()`): $L = 2/\ell_{\min}$, twice the reciprocal of
the smallest fitted lengthscale, so a wigglier objective (small
lengthscale, large $L$) gets smaller suppression balls, consistent with
the acquisition surface genuinely changing over shorter distances. If no
lengthscales can be extracted, $L$ defaults to 10 on the unit cube.

## The candidate pool

The acquisition function is not maximized with gradients. Each iteration
draws a fresh Latin hypercube candidate pool on the unit cube, scores
every candidate in one vectorized pass (a single batched GP prediction
per metric), and selects from the pool. The pool size adapts to the
problem: $500 \times d$ points, clamped to $[1000, 5000]$, with the
`candidate_pool` argument (default 2000) acting as a floor. A late-run
refinement inflates the pool by 50 percent (capped at 10,000) once the
iteration count exceeds 70 percent of the run's iteration budget,
`(budget - n_init) / q` rounded up (with `n_init` counted as the
evaluations the initial phase actually consumed), so the last 30 percent of BO
iterations search a denser pool regardless of how large the initial
design was. Derivative-free maximization over a space-filling pool is robust
to the multimodal, constraint-warped shape of the ECI surface and
composes directly with the sequential penalization rule above.

# 4. Multi-fidelity budgeting

## Fidelity levels and the simulator contract

`fidelity_levels` maps the labels `"low"`, `"med"`, `"high"` to
replication counts (default `c(low = 2000, med = 4000, high = 10000)`).
The names are accepted in any order: the vector is canonicalized to
ascending replication count on entry, so the first-named level is
always the *cheapest* one. The initial design is evaluated entirely at
that cheapest level, and the selection methods below fall back to it;
under the default values that is `"low"`. Each BO batch point then gets
its own fidelity decision.

How the chosen count reaches your simulator matters. BATON always passes
the fidelity *label* as the `fidelity` argument. If the simulator
declares an explicit `n_rep` argument in its formals, BATON additionally
passes the requested replication count, `n_rep =
fidelity_levels[[fidelity]]` (including dynamically escalated counts
under the `hybrid_staged` method), so the simulator can honor the exact
budget instead of re-deriving it from the label. Accepting `...` alone
does not opt in; this keeps wrappers that forward `...` to legacy
simulators working. The realized count is recorded in the history's
`n_rep` column, taken from the simulator's `n_rep` attribute when
supplied and from the requested count otherwise. One exception to the
opt-in rule: Stage 4 multi-seed verification (Section 5) always passes
`n_rep` to the simulator, so a simulator used with
`multi_seed_verify = TRUE` should accept `n_rep` explicitly or at least
absorb it via `...` (the documented simulator signature does).

## Budget accounting

The `budget` argument counts *evaluations*, not replications:
`eval_counter` increments once per simulator call, and the BO loop runs
while `eval_counter < budget`. Replications are tracked separately: the
cumulative sum of the history's `n_rep` column feeds the fidelity
methods' cost reasoning, against an approximate total simulation budget
of `budget * mean(fidelity_levels)`. Within a batch, the accounting uses
the *requested* counts of earlier in-batch picks (batch points are
conceptually simultaneous, which is what makes them independently
evaluable in parallel).

## The four selection methods

`fidelity_method` chooses among four strategies, dispatched per batch
point with that point's predicted feasibility probability, its
surrogate coefficient of variation (predictive sd over absolute
predictive mean), its acquisition value, the iteration number, and the
replication budget consumed so far.

**Threshold** (`"threshold"`, legacy) is a pure feasibility ladder:
$P(\text{feasible}) \ge 0.75$ gets high fidelity, $\ge 0.40$ gets
medium, everything else low. Simple and transparent: precision is spent
where the surrogate already believes a design is likely feasible.

**Staged** (`"staged"`) is iteration-scheduled. The first 30 iterations
run entirely at low fidelity. From iteration 31 to 100, a point is
promoted to high fidelity when it is both uncertain (CV above
`fidelity_cv_threshold`, default 0.05) and near the feasibility boundary
(feasibility probability inside `fidelity_prob_range`, default
$[0.2, 0.8]$); otherwise medium fidelity for promising points
($P \ge 0.4$) and low for the rest. After iteration 100, likely-feasible
points ($P > 0.6$) go straight to high fidelity.

**Hybrid staged** (`"hybrid_staged"`) layers three refinements on the
staged idea: the fidelity levels themselves grow with iteration
(`compute_dynamic_fidelity_levels()` scales the base levels up to
threefold in the late phase, with a safety factor that backs the scaling
off as the replication budget depletes); CV thresholds are
metric-specific (constraints get a threshold twice as tight as the
objective, on the argument that boundary precision is what feasibility
decisions consume) and are tightened further near the boundary and near
the current best design; and the CV that drives escalation switches from
the objective's to the worst constraint's when the point sits in the
boundary band. This is the most aggressive escalation schedule in the
package.

**Adaptive** (`"adaptive"`, the default) frames the choice as expected
information gain net of cost, in the spirit of cost-aware multi-fidelity
acquisition (Wu and Frazier 2016). `select_fidelity_adaptive()` first
builds a scalar value score for the candidate point,

$$
v = \frac{\tilde{\alpha}}{1 + \tilde{\alpha}} \cdot
w_u\!\left(\min\{1, \mathrm{CV}/0.3\}\right) \cdot
w_b\!\left(\sqrt{1 - |2P - 1|}\right),
\qquad \tilde{\alpha} = \frac{\max(\alpha, 0)}{|f^\ast|},
$$

the product of a saturating acquisition factor (the acquisition value
$\alpha$ normalized by the incumbent objective $f^\ast$, so the factor
is scale-free), an uncertainty factor that saturates at CV $= 0.3$, and
a boundary factor peaking at $P(\text{feasible}) = 0.5$. The affine
weights $w_u, w_b$ shift with the run's stage: before iteration 20 the
uncertainty factor dominates (exploration), between 20 and 60 the two
balance, after 60 the boundary factor dominates (exploitation).

What distinguishes the levels is a gain factor
$g_j = R_j / (R_j + R_{\mathrm{ref}})$, rescaled so the largest level
has gain 1, with $R_{\mathrm{ref}}$ the median replication count. This
is the shape of the posterior-variance reduction at the candidate when
the new observation arrives with Monte Carlo noise $\tau^2/R$: monotone
increasing and saturating in $R$, so higher fidelity buys more
information, with diminishing returns. Each level is then scored as

$$
\mathrm{net}_j = v\,g_j - \lambda\left(\tilde{c}_j^{\,\gamma} - 1\right),
\qquad \lambda = 0.05,
$$

where $\tilde{c}_j$ is the level's cost relative to the cheapest (by
default proportional to its replication count, overridable via
`fidelity_costs`) and the cost exponent

$$
\gamma = \min\{\,0.2 + 0.2\min(1, t/100) + 0.15\,\beta,\;\; 0.55\,\}
$$

increases with iteration $t$ and with the fraction $\beta$ of the
replication budget already consumed, so escalation gets harder to
justify as resources deplete. Levels whose replication count exceeds the
remaining budget are excluded outright, and an exploration override
forces low fidelity with probability $\max\{0.05, 0.5\,e^{-t/30}\}$. The
staged intent survives in three places: early iterations explore cheaply
because the exploration override fires often and $w_u$ favors the
uncertainty signal; mid-run, strong candidates near the boundary clear
the cost penalty and escalate; late iterations escalate only candidates
whose value score is proportionally larger, because $\gamma$ has risen
and the affordability guard protects the remaining budget.

Two consequences worth knowing. First, escalation is genuinely
value-driven: a candidate is promoted only when its scale-free value
score clears the cost penalty, so a smooth, fast-converging problem can
legitimately run entirely at low fidelity (the acquisition and boundary
signals fade as the surrogates converge), while a problem whose
incumbent is still moving escalates its strongest candidates to medium
or high. (Prior to v0.8 the value score had no fidelity dependence at
all, so the cheapest level always maximized value-per-cost and the
method never escalated; v0.8 fixed that.) Second, the demonstration
below uses `"threshold"` rather than the default: the threshold ladder
makes the routing transparent, since each pick's tier is a direct
function of its feasibility probability, whereas the adaptive score
depends on four interacting signals, and on a toy problem this small and
smooth the adaptive method concludes, correctly, that low fidelity
suffices throughout. The staged schedules likewise operate on iteration
scales (30 to 100 iterations) that a vignette-sized run never reaches.

## Demonstration: fidelity routing in a tiny run

A two-parameter toy calibration, small enough to knit quickly: minimize
expected enrollment subject to a power constraint, with deliberately tiny
replication counts. The simulator declares `n_rep`, so it receives the
exact requested count; the history records the fidelity label and count
for every evaluation.

```{r demo-fidelity-run}
toy_sim <- function(theta, fidelity = "low", seed = NULL, n_rep = NULL, ...) {
  p_true  <- plogis(6 * (theta$dose - 0.3) + 2 * (theta$cut - 0.5))
  en_true <- 25 + 20 * theta$dose + 10 * theta$cut
  p_hat   <- rbinom(1, n_rep, p_true) / n_rep
  en_hat  <- en_true + rnorm(1, 0, 5 / sqrt(n_rep))
  metrics <- c(power = p_hat, EN = en_hat)
  attr(metrics, "variance") <- c(power = max(p_hat * (1 - p_hat), 0.01) / n_rep,
                                 EN = 25 / n_rep)
  attr(metrics, "n_rep") <- n_rep
  metrics
}

fit <- bo_calibrate(
  sim_fun = toy_sim,
  bounds = list(dose = c(0, 1), cut = c(0, 1)),
  objective = "EN",
  constraints = list(power = c("ge", 0.6)),
  n_init = 12, q = 2, budget = 40, seed = 7,
  fidelity_levels = c(low = 100, med = 400, high = 1600),
  fidelity_method = "threshold",
  early_stop = list(enabled = FALSE),  # keep all iterations for display
  progress = FALSE
)

fit$status
with(fit$history, table(iter, fidelity))
```

Iteration 0 is the initial design, all at low fidelity by construction.
From iteration 1 on, each batch point is routed by its predicted
feasibility probability: exploratory picks in uncertain territory stay
at low fidelity, while picks the surrogates already believe feasible are
promoted to medium or high fidelity, so precision concentrates where
feasibility decisions are being made. (Early stopping is disabled here
only so the full table is displayed; a run this small would otherwise
converge and stop within a handful of iterations.) The recorded `n_rep`
column confirms the simulator received the requested counts:

```{r demo-fidelity-history}
tail(fit$history[, c("eval_id", "iter", "fidelity", "n_rep",
                     "objective", "feasible")], 8)
round(unlist(fit$best_theta), 3)
```

# 5. The loop end to end

With the components in place, one BO iteration of `bo_calibrate()` is,
in the actual execution order:

1. **Walltime check** (only if `max_walltime_s` is set): the cap is
   checked at the top of each iteration, never mid-batch, so a capped run
   always returns an internally consistent partial fit.
2. **Fit surrogates.** One GP per metric on the full history, warm-started
   from the previous iteration's fits, under a deterministic fit seed
   (Section 6). If warm-started fitting errors, retry cold.
3. **Compute the incumbent** $f^*$ (see below).
4. **Draw the candidate pool**: a fresh LHS of adaptive size on the unit
   cube.
5. **Score the pool once.** A single batched prediction per metric yields
   means and sds for every candidate; feasibility probabilities and
   acquisition scores (ECI, or the pre-feasible violation score) are
   computed vectorized from those predictions.
6. **Select the batch.** The top scorer if only one point remains in the
   budget; otherwise local penalization picks `q` spatially separated
   points.
7. **Choose fidelities.** Each batch point independently gets a fidelity
   from the configured method, using its cached predictions and the
   replication budget consumed so far.
8. **Evaluate and record.** The batch is evaluated (in parallel across
   forked workers when `options(BATON.cores = k)` is set on Unix), each
   evaluation under its own derived seed, and all rows are appended to
   the history in one bind.
9. **Callback and checkpoint hooks** (v0.7.0, both optional): the
   `callback` runs at the end of the iteration and can cancel the run
   cooperatively; `checkpoint_fun` receives a slim resumable snapshot
   every `checkpoint_every` iterations.
10. **Early-stopping checks** (below), then loop while evaluations
    remain.

After the loop, the surrogates are refit once on the final history
(skipped under `slim = TRUE`), diagnostics are generated, and, if
requested, Stage 4 multi-seed verification re-evaluates the selected
design at several independent seeds at high fidelity and stamps a
pass/fail verdict on the fit.

## Incumbent definitions

The incumbent $f^*$ anchors EI, so its definition matters under noise.
`incumbent_method` offers two:

- **BPM, best posterior mean (default).** Predict the objective
  surrogate at every feasible evaluated design and take the minimum
  posterior *mean*. Because the GP smooths Monte Carlo noise, BPM is
  robust to a lucky low-fidelity draw masquerading as a great design,
  the plug-in approach of Picheny et al. (2013).
- **BOI, best observed incumbent.** The minimum observed objective among
  feasible rows. Sensitive to noise, but independent of surrogate
  quality.

Both are computed preferentially over *high-fidelity* feasible
evaluations: once any high-fidelity evaluation exists in the history, the
incumbent restricts to feasible high-fidelity rows, falling back to
any-fidelity feasible rows when that restriction would leave nothing
(returning to "no incumbent" just because precision is still low would
wrongly flip the acquisition back into pre-feasible mode). If no feasible
design exists at any fidelity, the incumbent is $+\infty$ and the
pre-feasible acquisition of Section 3 is used. Note the returned
`best_theta`/`best_objective` of the fit are the best *observed*
feasible row when one exists, regardless of the incumbent method used
during the search; if the run ends with no feasible design at all,
`best_objective` is `NA` and `best_theta` is the best infeasible design,
flagged by a warning.

## Early stopping

Two convergence tests run at the end of every iteration (after a minimum
of 3 BO iterations), both configurable through `early_stop` and enabled
by default:

- **Improvement plateau.** Track the best feasible objective per
  iteration. If the relative improvement over the last `patience`
  iterations (default 5) falls below `threshold` (default $10^{-3}$, so
  0.1 percent), a counter increments; when it reaches `consecutive`
  (default 2) the run stops with status `"early_stopped"`. The
  comparison only happens within the feasible phase: before the first
  feasible point the tracked quantity is the best infeasible objective,
  a different quantity, and comparing across the transition could stop
  the run at the exact moment it first becomes feasible.
- **Acquisition flatline.** If the maximum acquisition value of the
  *selected batch* stays below a small level for 3 consecutive
  iterations, the run stops with status `"acq_flatline"`. Once a
  feasible incumbent exists the level is `acq_rel * |incumbent|`
  (default `acq_rel = 1e-3`), scaling the criterion with the objective's
  magnitude so it means the same thing for an objective of order 1 or
  order 100; before feasibility the level is `acq_rel` itself.

The fit's `status` field records why the run ended:
`"budget_exhausted"`, `"early_stopped"`, `"acq_flatline"`,
`"walltime"`, `"cancelled"`, or `"errored"`.

# 6. Implementation notes for reproducibility and scale

## Unit-hypercube internals

All modeling happens on $[0,1]^d$: `scale_to_unit()` /
`scale_from_unit()` map between raw parameter values and unit
coordinates, the LHS designs and candidate pools are drawn in unit
space, and GP lengthscales are unit-scale quantities. Raw values are
reconstructed only at the simulator boundary, where `integer_params` are
also rounded to the nearest integer before the call.

## Deduplication and aggregation via `theta_id`

Every design gets a deterministic string identifier: its unit
coordinates rounded to six decimals, fixed-notation formatted, and
joined (`theta_to_id()`; fixed notation per coordinate keeps the id
independent of session options and of other coordinates' magnitudes).
The id deduplicates LHS top-up draws against donor rows when
warm-starting from a previous run, and it defines the replicate groups
that the surrogate fitting aggregates (Section 2), so re-evaluations of
the same design reinforce one location instead of creating spurious
near-duplicate rows.

## The seeding architecture

Reproducibility is engineered around one base seed and deterministic
derivations from it:

- The run seed is `seed` (default 2025). With an explicit seed, the
  caller's RNG stream is restored on every exit path, so calling
  `bo_calibrate()` never advances `.Random.seed` in the calling
  environment. With `seed = NULL`, the run seed is drawn from the
  caller's stream with a single `sample.int()` draw (the stream advances
  by exactly one step, so repeated `seed = NULL` calls give independent
  replicates) and everything after is still rewound.
- **Every simulator call** runs under a derived seed, `seed + eval_id`,
  inside `run_seeded()`: the RNG state is saved, `set.seed()` applied,
  the call made, and the caller's state restored. The same derived seed
  is also passed to the simulator's `seed` argument so simulators
  managing their own streams stay reproducible.
- **Every GP fit** likewise runs under a derived seed (`seed + 10000 *
  iter + metric index` in-loop, `seed + 99999` for the final refit),
  because `km` and `mleHetGP` use randomized multi-starts whose outcome
  would otherwise depend on RNG stream position.

The consequence is a strong guarantee: results are bit-identical whether
evaluations and fits run serially or in parallel under
`options(BATON.cores = k)`, and regardless of the order forked workers
complete, because no random draw anywhere in the pipeline depends on
stream position.

## Parallel evaluation model

Parallelism is opt-in via `options(BATON.cores = k)` and uses fork-based
`parallel::mclapply()`, so it is effective on Unix (Linux, macOS) and
silently serial elsewhere (`effective_cores()` makes that decision in
one place). Three regions parallelize: the initial-design evaluations
(chunked, so a walltime cap can be checked at chunk boundaries without
losing throughput), each BO batch (its `q` points are independent by
construction), and the per-metric surrogate fits. Worker errors are
re-thrown to match serial behavior.

## Service controls (v0.7.0)

For long or unattended calibrations, `bo_calibrate()` exposes a set of
operational controls: a `status` field on every returned fit recording
why the run ended; `max_walltime_s` for an advisory walltime cap with a
consistent partial return; `on_error = "return_partial"` to convert
mid-run failures into a partial fit carrying every completed evaluation
and the error message; `callback` for per-iteration monitoring with
cooperative cancellation; `checkpoint_fun` / `checkpoint_every` for
periodic resumable snapshots (resume by passing a checkpoint's history
back as `initial_history`); and `slim = TRUE` for lightweight decision
objects that drop the large GP and diagnostic components while keeping
the history and the adoption gate. See `?bo_calibrate` for the full
contracts and `vignette("BATON-introduction", package = "BATON")` for
usage patterns.

# 7. What BATON does not do

Honest scoping, so you can decide quickly whether it fits your problem:

- **No equality constraints.** Constraints are one-sided (`"ge"` or
  `"le"`). An (approximate) equality target must be encoded as a pair of
  inequalities with a tolerance band, at the cost of a thin feasible
  region that is genuinely harder to search.
- **Box bounds only.** The design space is a product of intervals. Joint
  restrictions between parameters (say, an interim time that must precede
  the final analysis) should be handled by reparameterizing (optimize the
  gap rather than the second time point) or, less efficiently, by having
  the simulator return an infeasible-signaling metric value.
- **Single objective.** One metric is minimized. Multi-objective
  trade-offs are handled the classical way: minimize a weighted
  combination (the companion manuscript's weighted sample-size loss,
  spanning optimal, minimax, and admissible design philosophies, is
  exactly this recipe) and trace the frontier by re-running with
  different weights. A hypervolume-based acquisition (qEHVI) is future
  work; the internal `acq_qehvi` placeholder currently warns and falls
  back to ECI.
- **Gaussian-ish aggregate noise.** The GP machinery treats each
  observed metric as Gaussian around its true value, which is well
  justified for Monte Carlo *averages* over hundreds of replicates but
  degrades for extreme proportions at low fidelity (where the binomial
  is skewed) and for heavy-tailed summaries. Raising the low-fidelity
  replication count is the practical mitigation.

# 8. Citing BATON and references

If you use BATON in published work, please cite the methods manuscript;
this vignette paraphrases it and the package implements it:

> Young, A. M., Li, D., Hilsenbeck, S. G., Tayob, N., Chen, R., Yuan,
> Y., Bates, S., Kelly, E., Burns, R., Jhaveri, K., Tolaney, S. M.,
> Spears, P. A., Goetz, M. P., Davidson, N. E., Norton, L., Perou,
> C. M., Krop, I. E., Wolff, A. C., Winer, E. P., Carey, L. A., and
> Rashid, N. U. "BATON: Constrained Bayesian Optimization for
> Calibrating Adaptive Trials." Submitted.

Methodological references drawn on above:

- Binois, M., Gramacy, R. B., and Ludkovski, M. (2018). Practical
  heteroscedastic Gaussian process modeling for large simulation
  experiments. *Journal of Computational and Graphical Statistics*,
  27(4), 808-821. (The hetGP model and package.)
- Gardner, J. R., Kusner, M. J., Xu, Z., Weinberger, K. Q., and
  Cunningham, J. P. (2014). Bayesian optimization with inequality
  constraints. *ICML*.
- Gelbart, M. A., Snoek, J., and Adams, R. P. (2014). Bayesian
  optimization with unknown constraints. *UAI*.
- Gonzalez, J., Dai, Z., Hennig, P., and Lawrence, N. (2016). Batch
  Bayesian optimization via local penalization. *AISTATS*.
- Jones, D. R., Schonlau, M., and Welch, W. J. (1998). Efficient global
  optimization of expensive black-box functions. *Journal of Global
  Optimization*, 13, 455-492. (Expected improvement.)
- Picheny, V., Wagner, T., and Ginsbourger, D. (2013). A benchmark of
  kriging-based infill criteria for noisy optimization. *Structural and
  Multidisciplinary Optimization*, 48, 607-626. (Plug-in incumbents
  under noise.)
- Roustant, O., Ginsbourger, D., and Deville, Y. (2012). DiceKriging,
  DiceOptim: Two R packages for the analysis of computer experiments by
  kriging-based metamodeling and optimization. *Journal of Statistical
  Software*, 51(1), 1-55.
- Wu, J., and Frazier, P. (2016). The parallel knowledge gradient method
  for batch Bayesian optimization. *NeurIPS*. (Cost-aware multi-fidelity
  motivation.)

```{r session-info}
sessionInfo()
```
