---
title: "An End-to-End Calibration Case Study"
author: "BATON Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An End-to-End Calibration Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
# Suppress the one-time q-default migration note so chunk output stays clean.
options(BATON.q_warning_shown = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# Precomputed results for the heavy stages; see the note in the text.
cs <- readRDS("precompute/case_study_results.rds")
```

```{r load}
library(BATON)
```

This vignette carries one realistic trial calibration from problem statement
to final verified design, exercising the package's analysis helpers in the
order you would actually use them: build and test an evaluator, explore,
calibrate, verify, probe sensitivity, benchmark the method, quantify
reliability, and compare design philosophies. The two sibling vignettes
cover the ground on either side: `vignette("BATON-introduction")` teaches
the core workflow, and `vignette("BATON-methods")` explains the machinery.
This one is about the workflow around a real decision.

A note on reproducibility before we start. The expensive stages of this
case study (the main calibration, the benchmark, the reliability study, and
the philosophy comparison) are not executed when this document is knitted.
They were run by the included script
`vignettes/precompute/case-study-precompute.R`, which saves compact results
to `case_study_results.rds`; the vignette loads that file and displays the
stored output, and every code chunk shown for a precomputed stage is
verbatim what the script ran. Re-run the script to reproduce everything.
On this vignette's deliberately fast simulator the whole script takes well
under a minute; the pattern matters because with a realistic patient-level
simulator the same stages take hours to days, and you do not want a
document rebuild to trigger them. Cheap steps (the evaluator checks, the
quick exploration run, the verification arithmetic) do run live at knit
time.

# The Design Problem

A randomized two-arm phase II trial compares an experimental treatment
against control on a binary response endpoint. The control response rate
is assumed to be 25%, and the treatment is considered clinically
interesting if it lifts response to 45%. The sponsor wants one interim
analysis with both an efficacy boundary (stop early and declare success if
the interim result is overwhelming) and a binding futility boundary (stop
early and give up if the interim result is poor). The regulatory
constraints are conventional: one-sided type I error at most 0.025 and
power at least 0.85 under the alternative.

Four numbers fully specify a design in this family, and all four are
calibrated jointly:

| Parameter | Meaning |
|-----------|---------|
| `n_max` | Maximum total sample size (both arms, 1:1 allocation) |
| `t_interim` | Fraction of `n_max` enrolled at the interim look |
| `c_eff` | Efficacy boundary parameter: the final-analysis critical value; the interim efficacy boundary is `c_eff / sqrt(t)` (an O'Brien-Fleming-type shape) |
| `c_fut` | Interim futility boundary on the z scale (binding: cross it and the trial stops for futility) |

The objective is to minimize the expected total sample size under the
alternative, `EN`, subject to the two constraints. The clinical logic of
that choice: if the treatment works, early efficacy stopping gets the
answer with fewer patients randomized, shortens the trial, and accelerates
the next phase; if it does not work, the futility boundary limits patient
exposure, and the type I error constraint caps the risk of a false
positive. The trade-off being negotiated is real. A larger `n_max` buys
power but inflates every trial that runs to the end; an aggressive
efficacy boundary saves patients but spends type I error at the interim; a
tight futility boundary protects patients under the null but erodes power
under the alternative. No single parameter can be chosen sensibly in
isolation, which is what makes this a calibration problem rather than a
lookup.

# Building and Testing the Evaluator

The evaluator follows the canonical BATON contract
(`function(theta, fidelity, seed = NULL, n_rep = NULL, ...)`) and
simulates under both hypotheses in every call, because power must be
estimated under the alternative and type I error under the null. The
statistical content: patients are Bernoulli responders; at the interim,
with `n1` patients per arm, a pooled two-sample z statistic is compared to
the efficacy boundary `c_eff / sqrt(t)` and the futility boundary `c_fut`;
trials that continue enroll to `n_max` and reject if the final z statistic
exceeds `c_eff`. The chunk below is identical to the copy in the
precompute script.

```{r simulator}
p_control <- 0.25   # control-arm response rate (null and alternative)
p_treat   <- 0.45   # treatment-arm response rate under the alternative

gs_simulator <- function(theta, fidelity = c("low", "med", "high"),
                         seed = NULL, n_rep = NULL, ...) {
  fidelity <- match.arg(fidelity)
  if (is.null(n_rep)) {
    n_rep <- switch(fidelity, low = 2000, med = 4000, high = 10000)
  }
  if (!is.null(seed)) set.seed(seed)

  # Design parameters (theta arrives fractional; realize an integer design)
  n_arm  <- floor(round(theta$n_max) / 2)            # per-arm maximum
  n1     <- max(2L, round(theta$t_interim * n_arm))  # per-arm at interim
  n2     <- n_arm - n1                               # per-arm stage-2 increment
  t_frac <- n1 / n_arm                               # realized information fraction
  c_interim <- theta$c_eff / sqrt(t_frac)  # O'Brien-Fleming-type interim bound
  c_final   <- theta$c_eff                 # final-analysis critical value
  c_fut     <- theta$c_fut                 # binding interim futility bound

  # Pooled two-sample z statistic for a difference in proportions
  z_stat <- function(x_t, x_c, n) {
    pbar <- (x_t + x_c) / (2 * n)
    se <- sqrt(pmax(pbar * (1 - pbar) * 2 / n, 1e-12))
    ifelse(pbar <= 0 | pbar >= 1, 0, (x_t / n - x_c / n) / se)
  }

  run_hypothesis <- function(p_trt) {
    xt1 <- rbinom(n_rep, n1, p_trt)
    xc1 <- rbinom(n_rep, n1, p_control)
    z1  <- z_stat(xt1, xc1, n1)
    stop_eff <- z1 >= c_interim
    stop_fut <- !stop_eff & (z1 <= c_fut)
    continue <- !stop_eff & !stop_fut
    xt2 <- rbinom(n_rep, n2, p_trt)
    xc2 <- rbinom(n_rep, n2, p_control)
    z2  <- z_stat(xt1 + xt2, xc1 + xc2, n_arm)
    reject  <- stop_eff | (continue & z2 > c_final)
    n_total <- ifelse(continue, 2 * n_arm, 2 * n1)
    list(reject = reject, n_total = n_total)
  }

  alt  <- run_hypothesis(p_treat)    # alternative: treatment works
  null <- run_hypothesis(p_control)  # null: treatment equals control

  metrics <- c(power = mean(alt$reject),
               type1 = mean(null$reject),
               EN    = mean(alt$n_total))
  attr(metrics, "variance") <- c(power = var(alt$reject)  / n_rep,
                                 type1 = var(null$reject) / n_rep,
                                 EN    = var(alt$n_total) / n_rep)
  attr(metrics, "n_rep") <- n_rep
  metrics
}
```

Contract details worth pausing on. The `n_rep = NULL` default with the
`switch()` fallback means BATON's exact requested replication count is
used when BATON calls the evaluator, while manual calls (like the checks
below) get sensible tier defaults. The `variance` and `n_rep` attributes
turn on the heteroskedastic surrogate, which matters here: the Monte Carlo
variance of `EN` depends strongly on the design (a design that stops at
the interim 40% of the time has a bimodal sample-size distribution; one
that rarely stops concentrates at `n_max`).

Because each replicate is a couple of `rbinom()` draws, the whole tier is
vectorized and the variance of each mean is just `var(x) / n_rep`. When
one replicate is a genuine function call (a patient-level simulation with
accrual, dropout, and response modeling), you cannot vectorize, and
storing every replicate to compute the variance gets expensive at high
tiers. That is what `welford_mean_var()` is for: it runs the
one-replicate function `n_rep` times and returns the mean and the
variance of the mean in a single pass with O(1) memory. It computes
exactly the same quantities:

```{r welford-demo}
set.seed(42)
reps <- matrix(rnorm(5000 * 3), ncol = 3,
               dimnames = list(NULL, c("power", "type1", "EN")))
w <- welford_mean_var(function(i, data) data[i, ], n_samples = nrow(reps),
                      data = reps)
all.equal(w$mean, colMeans(reps))
all.equal(w$variance, apply(reps, 2, var) / nrow(reps))
```

Same numbers, never more than one replicate in memory. See
`vignette("variance-estimation")` for the algorithm and
`pool_welford_results()` for combining parallel chunks.

## Sanity Checks Before Any Optimization

Never hand an untested evaluator to an optimizer: a bug found now costs
seconds, the same bug found inside a calibration costs the whole run and,
worse, may not look like a bug in the aggregated output. Three checks, in
increasing order of strength.

**Shape and plausibility.** Call it once and look:

```{r sanity-shape}
test <- gs_simulator(list(n_max = 220, t_interim = 0.5, c_eff = 2.0, c_fut = 0),
                     fidelity = "low", seed = 1)
round(test, 4)
round(sqrt(attr(test, "variance")), 4)   # standard errors at the low tier
```

Named metrics, values in plausible ranges, variance and `n_rep` attributes
present. Power around 0.87 and type I error around 0.025 for a 220-patient
design with these boundaries is in the right neighborhood.

**An analytically checkable corner.** The design family contains the
classical fixed design as a degenerate case: with `t_interim = 1` the
interim and final analyses coincide and the decision rule collapses to a
single-look z test with critical value `c_eff` (the futility bound becomes
unreachable). That corner has a textbook answer, so simulate it and
compare against `power.prop.test()`:

```{r sanity-corner}
corner <- gs_simulator(list(n_max = 196, t_interim = 1, c_eff = qnorm(0.975),
                            c_fut = -10),
                       seed = 1, n_rep = 200000)
round(corner, 4)
power.prop.test(n = 98, p1 = 0.25, p2 = 0.45, sig.level = 0.025,
                alternative = "one.sided")$power
corner[["EN"]] == 196   # no early stopping possible, so EN must be exact
```

Simulated power 0.840 against the analytic approximation 0.841, type I
error within Monte Carlo error of the nominal 0.025, and `EN` exactly
equal to `n_max` because a one-look design cannot stop early. If any of
these disagreed materially, the boundary logic or the test statistic would
be wrong, and no amount of optimization downstream would be trustworthy.

**Noise arithmetic.** The constraint thresholds have to be compared
against the evaluator's precision. At the low tier (2,000 replicates), a
type I error near 0.025 is estimated with standard error
`sqrt(0.025 * 0.975 / 2000)`, about 0.0035, so single low-tier
evaluations cannot distinguish 0.025 from 0.032:

```{r sanity-noise}
round(c(low_2000  = sqrt(0.025 * 0.975 / 2000),
        med_4000  = sqrt(0.025 * 0.975 / 4000),
        high_10k  = sqrt(0.025 * 0.975 / 10000),
        check_200k = sqrt(0.025 * 0.975 / 200000)), 5)
```

Keep this table in mind: it explains most of what happens in the rest of
this vignette. The type I error constraint sits within about one low-tier
standard error of where the interesting designs live, which means observed
feasibility during the search is a noisy signal, and verification at far
higher replication is not optional.

# A Quick Exploration Run

The first calibration of a new problem should be sized to be thrown away:
its job is to find the problems with the problem, not to produce the
answer. The Quick profile (see the introduction vignette) is about
`5 x d` initial points and `10-15 x d` total budget. We run it live here
with a deliberately naive first guess at the bounds, including a `c_eff`
floor of 1.90 chosen only because it "looked wide enough":

```{r quick-run, message = FALSE}
bounds_quick <- list(
  n_max     = c(120, 200),
  t_interim = c(0.25, 0.80),
  c_eff     = c(1.90, 2.40),
  c_fut     = c(-0.50, 1.00)
)
constraints <- list(power = c("ge", 0.85), type1 = c("le", 0.025))

fit_quick <- bo_calibrate(
  sim_fun = gs_simulator,
  bounds = bounds_quick,
  objective = "EN",
  constraints = constraints,
  n_init = 20, q = 2, budget = 60, seed = 101,
  fidelity_levels = c(low = 1000, med = 4000, high = 10000),
  progress = FALSE
)
fit_quick$status
round(unlist(fit_quick$best_theta), 3)
fit_quick$best_objective
sum(fit_quick$history$feasible)
```

The run converged by acquisition flatline after
`r nrow(fit_quick$history)` of the 60 budgeted evaluations and reports a
remarkably cheap best design: about `r round(fit_quick$best_theta$n_max)`
patients with expected sample size around
`r round(fit_quick$best_objective)`. Before celebrating, look at where the
parameters landed. `c_eff` is sitting at
`r round(fit_quick$best_theta$c_eff, 3)`, essentially on its lower bound
of 1.90. A parameter pinned to a bound is always a finding: either the
box is too small (the optimizer wants something outside it) or the
constraint that should have stopped it did not bite. Re-checking the
winner at 200,000 replicates settles which:

```{r quick-recheck}
recheck <- gs_simulator(fit_quick$best_theta, seed = 999, n_rep = 200000)
round(recheck, 4)
```

The design that looked feasible during the search (observed power 0.875,
observed type I error 0.024 at 1,000 replicates) actually has power
`r round(recheck[["power"]], 3)` and type I error
`r round(recheck[["type1"]], 4)`. Both constraints are violated, the
type I error one badly. The mechanism is worth understanding because it is
generic, not a quirk of this problem: at 1,000 replicates the type I error
standard error is about 0.005, so among the many designs the search
visited near the constraint boundary, some drew lucky Monte Carlo samples
that understated their true error rate, and the optimizer, which is
explicitly hunting for the cheapest design that looks feasible, is a
machine for finding exactly those lucky draws. Optimization pressure turns
noise into bias.

The quick run has done its job, yielding three concrete adjustments:

1. **Raise the `c_eff` floor to 1.96.** This one comes with an analytic
   argument, not just an empirical one: a final critical value below
   `qnorm(0.975) = 1.96` cannot control one-sided type I error at 0.025
   even without an interim look, and adding an efficacy look only inflates
   it further. Every design with `c_eff < 1.96` is infeasible by
   construction, so the bound should exclude them and spend the budget
   elsewhere.
2. **Widen `n_max` upward.** The near-feasible mass piled up at the top of
   the `n_max` range, and the high-fidelity recheck says true power there
   is still short of 0.85. The feasible region evidently needs more
   patients than the naive box allowed; the balanced run uses
   `n_max` in [150, 280].
3. **Use a less noisy low tier.** With the type I error constraint one
   low-tier standard error from the action, 1,000-replicate evaluations
   are too blunt. The balanced run doubles the low tier to 2,000
   (standard error 0.0035) and keeps 10,000 for the high tier.

The general lesson: read the quick run's output as diagnostics, not as an
answer. Bound hits, near-boundary constraints, and a feasible fraction
near zero are all instructions for the next run.

# The Balanced Calibration Run

The production run uses the corrected bounds, the Balanced profile
(`n_init = 10 x d`, `budget` at the top of the `20-30 x d` range), the
multi-seed verification gate, and the long-run controls (checkpointing and
a walltime cap) that any run you care about should carry. This is the
heavy stage, so it is precomputed; the chunk below is verbatim from the
precompute script.

```{r bounds-cal}
bounds_cal <- list(
  n_max     = c(150, 280),   # widened after the quick run
  t_interim = c(0.25, 0.80),
  c_eff     = c(1.96, 2.40), # floor raised to the single-look critical value
  c_fut     = c(-0.50, 1.00)
)
constraints <- list(power = c("ge", 0.85), type1 = c("le", 0.025))
```

```{r balanced-run, eval = FALSE}
ckpt_dir <- tempdir()

fit_balanced <- bo_calibrate(
  sim_fun = gs_simulator,
  bounds = bounds_cal,
  objective = "EN",
  constraints = constraints,
  n_init = 40, q = 2, budget = 140, seed = 2026,
  fidelity_levels = c(low = 2000, med = 4000, high = 10000),
  multi_seed_verify = TRUE, multi_seed_n = 5,
  max_walltime_s = 4 * 3600,
  checkpoint_fun = function(snap) {
    saveRDS(snap, file.path(ckpt_dir, "balanced-checkpoint.rds"))
  },
  checkpoint_every = 5,
  progress = FALSE
)
```

```{r balanced-load}
fit_balanced <- cs$balanced   # stored result of exactly the call above
fit_balanced$status
round(unlist(fit_balanced$best_theta), 3)
fit_balanced$best_objective
```

## Reading the Run

```{r balanced-history}
nrow(fit_balanced$history)
table(fit_balanced$history$fidelity)
sum(fit_balanced$history$feasible)
```

Three observations, each worth a comment.

First, the run stopped at `r nrow(fit_balanced$history)` of 140 budgeted
evaluations with `status = "acq_flatline"`: the acquisition function
collapsed because, given everything the surrogates had seen, no candidate
offered meaningful expected improvement. That is a healthy ending, and on
this problem it is also believable, because the four-dimensional space is
smooth and the surrogates learn it quickly. Budget you did not need is
budget saved.

Second, every evaluation ran at the low tier. The adaptive fidelity
selector promotes evaluations when higher replication buys information
the search actually needs, and on this fast, smooth problem it decided
low-tier information was sufficient throughout. This is a legitimate
allocation, but it has a consequence you must not forget: nothing in the
optimization history has been measured to better than a 0.0035 standard
error on type I error, so the verification stage is carrying all of the
high-precision burden. On harder problems, or with `fidelity_costs`
reflecting a genuinely expensive high tier, you will see mixed
allocations.

Third, `r sum(fit_balanced$history$feasible)` of the evaluations were
observed-feasible. The feasible region is a thin shell here (the
constraints pin both power and type I error near their thresholds at the
optimum), so most of the budget went to mapping the boundary, which is
where the information about the constrained optimum lives.

The incumbent trace shows the classic trajectory, big early gains and a
long refinement tail:

```{r incumbent-trace, fig.alt="Step plot of the best feasible expected sample size by BO iteration, dropping steeply in early iterations and flattening near 168"}
h <- fit_balanced$history
inc <- vapply(split(seq_len(nrow(h)), h$iter), function(ix) {
  seen <- h[seq_len(max(ix)), ]
  feas <- seen[seen$feasible, ]
  if (nrow(feas) > 0) min(feas$objective) else NA_real_
}, numeric(1))
plot(as.integer(names(inc)), inc, type = "s", lwd = 2,
     xlab = "BO iteration (0 = initial design)",
     ylab = "Best feasible EN so far")
```

`check_convergence()` applies the improvement-stagnation test to any
history, which is useful when you are deciding whether to extend a run or
in multi-stage workflows:

```{r check-convergence}
check_convergence(fit_balanced$history, objective = "EN", patience = 5,
                  improvement_threshold = 1e-3)
```

Note the two convergence lenses disagree slightly, and honestly: the
improvement test says "not stagnant" because the incumbent still crept
from 169.0 to 168.0 inside the last patience window, while the
acquisition flatline that actually ended the run says "no candidate looks
worth evaluating". Both are true. The acquisition view is forward-looking
(the model believes remaining gains are below the noise floor); the
improvement view is backward-looking. When they disagree in this
direction, the run ended at the right time.

With `ggplot2` installed, the two built-in pictures of the run show the
geometry. The frontier plot makes the thin feasible shell visible, and the
trade-off plot shows the optimizer pushing the low-EN designs toward, but
not across, the constraint boundaries:

```{r frontier, eval = has_ggplot2, fig.alt="Scatter plot of power against expected sample size for all evaluations, colored by feasibility; feasible points form a band above the power threshold"}
plot_feasible_frontier(fit_balanced)
```

```{r tradeoff, eval = has_ggplot2, fig.alt="Scatter plot of power against type I error colored by expected sample size; cheap designs cluster near both constraint boundaries"}
plot_tradeoff_surfaces(fit_balanced)
```

The best design itself: about
`r round(fit_balanced$best_theta$n_max)` patients maximum, interim at
`r round(100 * fit_balanced$best_theta$t_interim)`% of enrollment, final
critical value `r round(fit_balanced$best_theta$c_eff, 3)` (a whisker
above 1.96: the optimizer spent almost exactly the alpha the interim look
freed up), and a mild futility bound at z =
`r round(fit_balanced$best_theta$c_fut, 2)`. Expected sample size
`r round(fit_balanced$best_objective, 1)` as observed during the search.
Whether to believe those numbers is the next section's business.

# The Trust Pipeline

A calibrated design is a claim, not a fact. The claim was assembled from
low-fidelity Monte Carlo estimates by a procedure that actively seeks
favorable noise, so before anything is adopted it goes through three
steps: the built-in multi-seed verification verdict, a manual
high-replication re-check, and integer rounding with re-verification.

## Step 1: The Multi-Seed Verdict

Because the balanced run set `multi_seed_verify = TRUE`, BATON already
re-evaluated the winning design at five independent seeds at the high
tier (10,000 replicates each) and attached a verdict:

```{r verdict}
fit_balanced$verdict
round(unlist(fit_balanced$multi_seed_summary[
  c("n_seeds", "n_rep", "strict_feas_count", "power_mean", "power_sd",
    "type1_mean", "type1_sd", "EN_mean", "EN_sd")]), 4)
```

The verdict is `MULTI_SEED_FAIL`: only
`r fit_balanced$multi_seed_summary$strict_feas_count` of 5 verification
seeds satisfied every constraint. The summary shows why. Across seeds,
type I error averages `r round(fit_balanced$multi_seed_summary$type1_mean, 4)`
(fine), but power averages
`r round(fit_balanced$multi_seed_summary$power_mean, 4)`, about 0.2
percentage points short of the 0.85 floor, with a cross-seed standard
deviation of only
`r round(fit_balanced$multi_seed_summary$power_sd, 4)`. That is not a
fluke of one seed; the design's true power is just under the threshold.

This outcome should be expected, not alarming, and it is worth being
direct about why. The constrained optimum of this problem sits exactly on
the power boundary (any power above 0.85 could be traded for fewer
patients, so the optimizer walks downhill until the constraint is taut).
A design whose true power is exactly 0.85 passes an all-seeds-strict
verification only rarely, and low-fidelity selection bias (the same
mechanism the quick run exposed) tends to leave the raw calibrated design
a hair on the wrong side. The verdict did precisely its job: it stopped a
marginally infeasible design from being adopted silently. What it costs
to fix is the next step's question.

## Step 2: Manual High-Fidelity Re-Check

First, confirm the diagnosis at simulation ground truth. 200,000
replicates give standard errors of about 0.0008 for power and 0.00035 for
type I error:

```{r manual-recheck}
recheck_bal <- gs_simulator(fit_balanced$best_theta, seed = 4001,
                            n_rep = 200000)
round(recheck_bal, 4)
```

True power `r round(recheck_bal[["power"]], 4)`, true type I error
`r round(recheck_bal[["type1"]], 4)`: the type I error constraint holds
with a little room, the power constraint misses by about 0.3 percentage
points. The remedy is to buy the missing power with the cheapest currency
available. The sensitivity analysis in the next section will confirm this
quantitatively, but the candidate is obvious: increase `n_max` slightly,
holding the timing and boundaries at their calibrated values.

## Step 3: Round to a Protocol Design and Re-Verify

A protocol cannot enroll 202.8 patients. Rounding gives a maximum of 101
per arm with the interim after 61 per arm; while we are at it, scan a few
neighboring values of `n_max` at high replication to find the smallest
design that clears the power floor with margin rather than by luck:

```{r nmax-scan}
theta_cal <- fit_balanced$best_theta
scan <- do.call(rbind, lapply(seq(202, 222, by = 4), function(N) {
  th <- theta_cal
  th$n_max <- N
  m <- gs_simulator(th, seed = 4002, n_rep = 200000)
  data.frame(n_max = N, power = round(m[["power"]], 4),
             type1 = round(m[["type1"]], 4), EN = round(m[["EN"]], 1))
}))
scan
```

The scan is monotone and quiet: each 4 added patients buys roughly half a
percentage point of power at a cost of about 3 expected patients, and
type I error drifts slightly downward (larger trials make the normal
approximation of the test statistic better behaved). The 0.85 floor is
first cleared at `n_max = 206`, but only by 0.3 percentage points, which
is the kind of margin that future re-estimation or a slightly different
simulation model could erase. We adopt `n_max = 210`, whose estimated
power of 0.860 holds the floor with about one percentage point to spare.
That choice is a judgement call, and it is worth making it explicitly and
recording the rule: adopt the smallest design whose estimated power
clears the constraint by more than what you consider erasable margin,
here 0.005.

The final protocol design, restated in enrollable integers:

```{r final-design}
theta_final <- theta_cal
theta_final$n_max <- 210
n_arm  <- 105                                     # per arm
n1     <- round(theta_final$t_interim * n_arm)    # interim at 64 per arm
c(n_arm = n_arm, interim_per_arm = n1,
  realized_t = round(n1 / n_arm, 3),
  interim_eff_bound = round(theta_final$c_eff / sqrt(n1 / n_arm), 3),
  final_bound = round(theta_final$c_eff, 3),
  futility_bound = round(theta_final$c_fut, 3))
```

And the re-verification, in the same spirit as the built-in gate: five
fresh seeds, high replication, all constraints checked at each seed:

```{r final-verify}
verify <- do.call(rbind, lapply(1:5, function(s) {
  m <- gs_simulator(theta_final, seed = 5000 + s, n_rep = 100000)
  data.frame(seed = 5000 + s, power = round(m[["power"]], 4),
             type1 = round(m[["type1"]], 4), EN = round(m[["EN"]], 1))
}))
verify
all(verify$power >= 0.85 & verify$type1 <= 0.025)
```

Every seed passes, with power between
`r min(verify$power)` and `r max(verify$power)` and type I error at most
`r max(verify$type1)`. This design, `n_max = 210` with the calibrated
timing and boundaries, is what goes in the protocol, and it is the design
all remaining sections analyze.

# Sensitivity Analysis

The calibration found a point; sensitivity analysis maps the terrain
around it, which is what turns an optimizer's output into design
understanding. Three helpers, all driven by the fitted GP surrogates (no
new simulation): global Sobol indices, local gradients at the optimum,
and the covariance of local effects across the space. These were computed
by the precompute script against the balanced fit's surrogates:

```{r sens-code, eval = FALSE}
set.seed(303)
sens <- list(
  sobol_EN    = sa_sobol(fit_balanced$surrogates, bounds_cal,
                         outcome = "EN", n_mc = 4000),
  sobol_power = sa_sobol(fit_balanced$surrogates, bounds_cal,
                         outcome = "power", n_mc = 4000),
  grad_EN     = sa_gradients(fit_balanced$surrogates, fit_balanced$best_theta,
                             bounds_cal, outcome = "EN"),
  grad_power  = sa_gradients(fit_balanced$surrogates, fit_balanced$best_theta,
                             bounds_cal, outcome = "power"),
  cov_EN      = cov_effects(fit_balanced$surrogates, bounds_cal,
                            outcome = "EN", n_mc = 500)
)
```

(One storage note, because it affects how you run this yourself: a fit
saved with `slim = TRUE`, or stripped of its `surrogates` for compact
storage as this vignette's precompute script does, cannot feed these
functions. Compute sensitivity while the full fit is in memory, or store
the full fit. The script computes the tables first and strips afterward.)

## What Matters Globally

```{r sobol}
cs$sens$sobol_EN
cs$sens$sobol_power
```

```{r sobol-plot, eval = has_ggplot2, fig.alt="Bar chart of first-order Sobol indices for expected sample size; n_max dominates with an index near 0.85"}
plot_sobol_indices(cs$sens$sobol_EN)
```

The Sobol indices give first-order fractions of variance across the whole
box. For `EN`, `n_max` explains about 85% of the variance and
`t_interim` about 6%, with the boundary parameters contributing almost
nothing; averaged over the design space, expected sample size is mostly
just trial size. For `power` the picture is similar but with `c_eff`
taking a real share (about 18%): where you set the final critical value
matters for power, but which z-score triggers futility barely does within
this box. Note what the near-zero `c_fut` indices do and do not say: they
say EN under the alternative and power are insensitive to the futility
bound in this range (futility stops are rare when the treatment works);
they say nothing about EN under the null, which is where futility earns
its keep and which this calibration did not target.

## What Matters at the Optimum

```{r gradients}
cs$sens$grad_EN
cs$sens$grad_power
```

Local gradients answer the design-negotiation questions that always come
after calibration. Reading the point estimates (the `sd` column here is
very conservative: it treats the two finite-difference endpoints as
independent posterior draws, so treat magnitudes and signs qualitatively):

- `dEN/dn_max` is about 0.79: each patient added to the cap adds about
  0.8 expected patients, because most trials run past the interim. This
  is the price of the power margin bought in the trust pipeline (8 extra
  `n_max` cost about 5 expected patients, exactly what the scan showed).
- `dpower/dt_interim` is about 0.05 per unit of `t_interim`, so moving
  the interim look by 5% of enrollment changes power by roughly 0.003,
  within Monte Carlo noise. Power is locally flat in interim timing:
  **operational flexibility exists**. If site logistics need the interim
  a few weeks earlier or later than the 61% point, the power case does
  not move. The `EN` gradient in `t_interim` is larger (about -44 per
  unit, or 2 expected patients per 5% shift), so timing is an efficiency
  question, not a validity question.
- `dpower/dc_eff` is about -0.16: raising the final critical value by
  0.1 costs 1.6 percentage points of power. Together with the near-zero
  Sobol index of `c_fut`, this confirms the efficacy boundary is the
  load-bearing decision threshold and the futility bound is, for these
  operating characteristics, close to a free parameter.

## How Effects Covary

```{r cov-effects}
round(cs$sens$cov_EN, 1)
```

```{r cov-plot, eval = has_ggplot2, fig.alt="Heatmap of the covariance matrix of local EN gradients; the t_interim diagonal entry dominates and t_interim covaries negatively with c_eff"}
plot_case_covariance(cs$sens$cov_EN)
```

`cov_effects()` samples local gradients of `EN` at 500 random points and
reports their covariance: it measures how much each parameter's effect
varies across the space, and which effects move together. The dominant
entry is the `t_interim` variance (about 5,200): the effect of interim
timing on `EN` changes sign and magnitude dramatically depending on where
you are, which is why the global Sobol index for `t_interim` is small
even though its local gradient at the optimum is not. The negative
`t_interim`-`c_eff` covariance says those two interact: with a stricter
final boundary, delaying the interim has a different EN consequence than
with a lax one. The practical reading: statements about interim timing
are only meaningful jointly with the boundary choice, so if the boundary
is renegotiated late (a regulator preferring a more conservative shape,
say), the timing should be re-examined too rather than carried over.

# Benchmarking the Calibration Method

Was Bayesian optimization worth the machinery on this problem?
`benchmark_methods()` runs the same calibration with competing strategies
under a common evaluation harness. Here BO (3 replicate runs at the
balanced settings) is compared against random search (3 runs of 100
LHS-sampled designs, each evaluated at the high tier) and a grid (3
points per axis, 81 designs at the high tier). Precomputed:

```{r bench-code, eval = FALSE}
bench <- benchmark_methods(
  sim_fun = gs_simulator,
  bounds = bounds_cal,
  objective = "EN",
  constraints = constraints,
  strategies = c("bo", "random", "grid"),
  bo_args = list(n_init = 40, q = 2, budget = 140, seeds = 1:3,
                 fidelity_levels = c(low = 2000, med = 4000, high = 10000),
                 slim = TRUE),
  random_args = list(n_samples = 100, seeds = 1:3),
  grid_args = list(resolution = 3),
  progress = FALSE
)
bench_summary <- summarise_benchmark(bench)
```

```{r bench-summary}
as.data.frame(cs$bench_summary[, c("strategy", "runs", "feasible_rate",
                                   "objective_mean", "objective_sd",
                                   "sims_mean")])
```

```{r bench-plot, eval = has_ggplot2, fig.alt="Scatter plot of mean best objective against mean simulation replicates per strategy; BO sits in the low-cost low-EN corner"}
plot_benchmark_efficiency(cs$bench)
```

What the data actually show, including the caveats:

- **BO wins on both axes here.** Mean best feasible `EN` of about
  `r round(cs$bench_summary$objective_mean[cs$bench_summary$strategy == "bo"])`
  using about
  `r round(cs$bench_summary$sims_mean[cs$bench_summary$strategy == "bo"] / 1000)`k
  simulator replicates per run, against about
  `r round(cs$bench_summary$objective_mean[cs$bench_summary$strategy == "random"])`
  for random search at 1,000k replicates and
  `r round(cs$bench_summary$objective_mean[cs$bench_summary$strategy == "grid"])`
  for the grid at 810k. Random search with roughly eight times the
  simulation budget lands about 24 expected patients worse; the coarse
  grid is not competitive, and a grid fine enough to be competitive in
  four dimensions would need orders of magnitude more points.
- **The comparison is not entirely apples to apples, in a direction that
  flatters BO.** The baselines evaluated every design at 10,000
  replicates, so their observed-best values are solid, while BO's
  observed best comes from 2,000-replicate evaluations and is
  optimistically biased by the selection effect this vignette keeps
  running into (the trust pipeline showed the raw BO winner really sits
  at the boundary). The honest headline is not "BO finds EN 167 and
  random finds 190"; it is "after verification and margin, BO's design
  costs about 174 expected patients, still roughly 11-16 patients better
  than the best unverified random-search design, on an eighth of the
  simulation budget".
- **The variance column matters.** Random search's best-found value
  swings by 11 patients across seeds; BO's spread is 2.5. Reproducibility
  of the answer across reruns is itself a benefit.

# Constraint Reliability Across Seeds

The benchmark asked how good the answers look; the reliability study asks
whether they are true. `estimate_constraint_reliability()` repeats the
whole calibration across seeds and validates each winning design at
200,000 replicates, recording how often the calibrated design actually
satisfies its constraints:

```{r rel-code, eval = FALSE}
rel <- estimate_constraint_reliability(
  sim_fun = gs_simulator,
  bounds = bounds_cal,
  objective = "EN",
  constraints = constraints,
  strategies = c("bo", "random"),
  calibration_seeds = 1:5,
  validation_reps = 200000,
  bo_args = list(n_init = 20, q = 2, budget = 60,
                 fidelity_levels = c(low = 2000, med = 4000, high = 10000),
                 slim = TRUE),
  random_args = list(n_samples = 60),
  progress = FALSE
)
```

```{r rel-summary}
as.data.frame(cs$rel$summary)
as.data.frame(cs$rel$runs[, c("strategy", "run_id", "calibration_feasible",
                              "validation_feasible", "validation_objective")])
```

```{r rel-plot, eval = has_ggplot2, fig.alt="Bar chart of validation feasibility rates; random search at 100 percent, Bayesian optimization at 0 percent"}
plot_constraint_reliability(cs$rel)
```

This is the most instructive table in the vignette, and it deserves a
completely honest reading, because at first glance it looks like an
indictment of BO. Every one of the five BO calibrations produced a design
that was observed-feasible during calibration and **failed** validation at
200,000 replicates; every random-search design validated cleanly. The
resolution of the apparent paradox is in the objective column: the BO
designs that "fail" have validated `EN` around
`r round(cs$rel$summary$objective_mean[cs$rel$summary$strategy == "bo"])`
and are failing because their true power is a fraction of a percentage
point under 0.85, exactly like the balanced run's raw winner. The random
designs "pass" because random search almost never lands within a hair of
the constraint boundary; its winners sit comfortably inside the feasible
region, about
`r round(cs$rel$summary$objective_mean[cs$rel$summary$strategy == "random"] - cs$rel$summary$objective_mean[cs$rel$summary$strategy == "bo"])`
expected patients more expensive.

So the reliability study is not measuring "which method is right"; it is
measuring distance-to-boundary, and it quantifies a structural fact about
constrained calibration under noise: **an effective optimizer will ride
the constraint boundary, and boundary-riding designs will not validate
strictly without an explicit margin.** That is why the trust pipeline is
a pipeline: verdict, re-check, margin, re-verify. The 0% raw reliability
here is precisely the failure the margin step (adopting `n_max = 210`
over 202) repaired, at a verified cost of about 5 expected patients. If
you want raw calibrations to validate more often without a manual margin
step, calibrate against buffered thresholds (for example, power at least
0.86 against a protocol floor of 0.85) and verify against the protocol
values; the buffer plays the same role as the margin, chosen in advance.

# Calibrating a Family of Designs: Philosophies

A single constraint set is one philosophy about the trial. Stakeholder
discussions usually involve several: a sponsor pushing for the smallest
viable trial, a statistician defending power, a safety board wanting
conservatism. `bo_calibrate_philosophies()` calibrates a named family of
objective/constraint sets in one batch, ordering them so that designated
donor philosophies run first and later ones warm-start from the donors'
calibrated designs, with the multi-seed gate applied to every member.
Here the `balanced` philosophy (the one calibrated above) acts as donor,
and a `conservative` (power at least 0.90) and an `aggressive` (power at
least 0.80) variant warm-start from it:

```{r phil-code, eval = FALSE}
philosophies <- list(
  balanced     = list(objective = "EN",
                      constraints = list(power = c("ge", 0.85),
                                         type1 = c("le", 0.025))),
  conservative = list(objective = "EN",
                      constraints = list(power = c("ge", 0.90),
                                         type1 = c("le", 0.025))),
  aggressive   = list(objective = "EN",
                      constraints = list(power = c("ge", 0.80),
                                         type1 = c("le", 0.025)))
)

phil <- bo_calibrate_philosophies(
  scenario_id = "resp2arm",
  sim_fun = gs_simulator,
  bounds = bounds_cal,
  philosophies = philosophies,
  multi_seed_verify = TRUE, multi_seed_n = 5, multi_seed_strict = FALSE,
  n_init = 30, q = 2, budget = 100, seed = 11,
  fidelity_levels = c(low = 2000, med = 4000, high = 10000),
  slim = TRUE, progress = FALSE
)
```

The batch returns a manifest recording the dependency roles, warm-start
donors, verdicts, and runtimes:

```{r phil-manifest}
cs$phil_manifest[, c("philosophy", "dependency_role", "verdict",
                     "objective_value", "warmstart_donors")]
```

And the per-philosophy designs with their multi-seed-verified operating
characteristics (assembled by the precompute script from each fit's
`multi_seed_summary`):

```{r phil-table}
cs$phil_table
```

Reading the family honestly:

- **The power ladder has a visible price.** Verified expected sample size
  runs about 150 (aggressive, power 0.82) to 165 (balanced, power 0.85)
  to 188 (conservative, power 0.90): each 5 percentage points of power
  costs roughly 15 to 23 expected patients in this scenario. Putting that
  table in front of stakeholders is usually worth more than any single
  calibrated design, because it converts a philosophical argument into a
  priced menu.
- **Every verdict is a warning, and correctly so.** All three designs
  ride their respective constraint boundaries, and the verified operating
  characteristics show the same marginal violations the balanced run
  exhibited (type I error a shade over 0.025 for the two philosophies
  whose `c_eff` landed on the 1.96 floor; conservative's power a shade
  under 0.90). With `multi_seed_strict = FALSE` these surface as
  `MULTI_SEED_WARN` rather than `FAIL`, which keeps the donor usable for
  warm-starting while still flagging every member for the same
  margin-and-re-verify treatment Section
  "The Trust Pipeline" applied to the balanced design. With the default
  `multi_seed_strict = TRUE`, a failed donor is excluded from
  warm-starting recipients, a conservatism you want when a donor might
  be genuinely broken rather than marginally boundary-riding.
- **Warm-starting is about economy.** The manifest shows both recipients
  seeded from `balanced`; each recipient's search starts from a
  known-good region of a closely related problem rather than from
  scratch. On this toy problem the savings are seconds; on a
  patient-level simulator, donors can save a recipient a large fraction
  of its budget, and the dependency ordering (donors first) is exactly
  the workflow for calibrating a protocol's design family overnight.

Use this tool when the deliverable is a family (a design options table
for a protocol meeting, a sensitivity-of-design-to-philosophy appendix),
not just one design. For cross-scenario warm-starting and the donor
mechanics in more depth, see `vignette("v04-hardening")`.

# Hygiene for Long Runs

Everything in this case study ran in seconds because the simulator is a
toy. The workflow is designed for the real case, where a single balanced
calibration is an overnight job, and the balanced run above already
demonstrated the pattern to copy:

- **Checkpointing**: `checkpoint_fun` receives a slim, resumable snapshot
  every `checkpoint_every` iterations. If the job dies, resume by feeding
  the checkpoint's history back in:

```{r resume-pattern, eval = FALSE}
ck <- readRDS(file.path(ckpt_dir, "balanced-checkpoint.rds"))
fit_resumed <- bo_calibrate(
  sim_fun = gs_simulator, bounds = bounds_cal, objective = "EN",
  constraints = constraints,
  initial_history = ck$history,
  n_init = nrow(ck$history), budget = 140, seed = 2027,
  fidelity_levels = c(low = 2000, med = 4000, high = 10000)
)
```

- **Walltime capping**: `max_walltime_s = 4 * 3600` makes the run stop
  gracefully (with `status = "walltime"` and a usable partial fit) before
  a scheduler kills it ungracefully. Pair it with
  `on_error = "return_partial"` on flaky infrastructure so a single
  failed simulator call costs one evaluation, not the run.
- **`slim = TRUE`** returns a lightweight fit without surrogates and
  diagnostics, skipping the final surrogate refit. The precompute script
  uses it for every fit whose surrogates nothing downstream needs (the
  benchmark, reliability, and philosophy sub-runs), but deliberately
  **not** for the balanced fit, because Section "Sensitivity Analysis"
  needs the surrogates; there the script computes the sensitivity tables
  first and strips the fit before saving. Decide `slim` by asking what
  will consume the fit.
- **Batch parallelism**: with `q > 1`, setting `options(BATON.cores = k)`
  evaluates the points within each batch in parallel on Unix/macOS. This
  multiplies with any parallelism inside your evaluator (see
  `pool_welford_results()` in the introduction vignette), so budget cores
  across the two levels deliberately.

# The Payoff

The naive alternative to all of the above is the fixed design: no interim
look, sample size straight from `power.prop.test()`. That needs 101 per
arm (202 patients total), every one of them enrolled every time. Its
operating characteristics under the same simulation model, and the final
calibrated design's, computed live:

```{r payoff}
n_fixed <- 2 * ceiling(power.prop.test(power = 0.85, p1 = 0.25, p2 = 0.45,
                                       sig.level = 0.025,
                                       alternative = "one.sided")$n)
fixed_oc <- gs_simulator(list(n_max = n_fixed, t_interim = 1,
                              c_eff = qnorm(0.975), c_fut = -10),
                         seed = 4003, n_rep = 200000)
final_oc <- gs_simulator(theta_final, seed = 4004, n_rep = 200000)

summary_tbl <- data.frame(
  design = c("Naive fixed", "Calibrated GS (verified)"),
  n_max = c(n_fixed, 210),
  power = round(c(fixed_oc[["power"]], final_oc[["power"]]), 3),
  type1 = round(c(fixed_oc[["type1"]], final_oc[["type1"]]), 4),
  EN = round(c(fixed_oc[["EN"]], final_oc[["EN"]]), 1)
)
summary_tbl
```

The calibrated group-sequential design holds both constraints with
verified margin and saves about
`r round(fixed_oc[["EN"]] - final_oc[["EN"]])` patients in expectation
under the alternative, a
`r round(100 * (fixed_oc[["EN"]] - final_oc[["EN"]]) / fixed_oc[["EN"]])`%
reduction, in exchange for a maximum enrollment 8 patients higher than
the fixed trial's if the interim does not stop. The mechanics of the
saving are transparent: about
`r round(100 * (210 - final_oc[["EN"]]) / (210 - 2 * 64))`% of trials
under the alternative stop at the interim with 128 patients enrolled,
almost all of them for efficacy. If the treatment works, roughly four
trials in nine finish 82 patients early with the answer in hand.

Just as important is what the calibration did *not* cost. The
calibration path itself (the quick run, the balanced run, and the
built-in verification gate) consumed about 220,000 simulated trials per
hypothesis, and the high-precision trust-pipeline checks (the 200,000-rep
re-checks, the `n_max` scan, and the final five-seed verification) added
roughly another 2.3 million. The benchmark and reliability studies cost
millions more, but those are method-evaluation overhead for this
vignette, not something a practitioner runs for each design. Even the
everything-included total is far below what a competitive grid would
spend on one constraint set: 10 points per axis in four dimensions at
10,000 replicates is 100 million trials per hypothesis, and the grid
still could not resolve the constraint boundary this problem's optimum
sits on.

The pattern of this case study transfers directly to expensive
simulators; only the evaluator's body and the wall-clock scale change.
Quick run to debug the problem statement, balanced run with verification
to get a candidate, trust pipeline to make it adoptable, sensitivity to
understand it, and the batch tools when the deliverable is a family of
designs rather than a point.

# Session Info

```{r session-info}
sessionInfo()
```
