---
title: "Introduction to BATON: Bayesian Optimization for Clinical Trial Calibration"
author: "BATON Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BATON}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
# Suppress the one-time q-default migration note so chunk output stays clean.
options(BATON.q_warning_shown = TRUE)
```

# What Problem BATON Solves

Designing an adaptive clinical trial means choosing design parameters, for
example the maximum sample size, the timing of an interim look, and the
decision thresholds applied at each analysis, so that the trial achieves
adequate power while controlling type I error, and at the same time
minimizes a quantity you care about such as the expected sample size. These
performance measures are collectively called operating characteristics. For
adaptive designs they are usually impossible to compute analytically: the
only way to know the power of a two-stage design with a futility rule is to
simulate thousands of trial replicates and count how often the trial rejects.
Calibration is therefore a constrained optimization problem where every
function evaluation is expensive and every observed value is noisy, because
it is a Monte Carlo estimate rather than an exact number.

Grid search, the traditional approach, fails on both counts. A grid fine
enough to locate a good design in four parameters at ten points per axis
needs 10,000 simulator calls, almost all of them spent in regions a few
early evaluations would have ruled out. And because each grid value is
noisy, the apparent best cell may just be the luckiest one. BATON replaces
the grid with constrained Bayesian optimization: it fits a Gaussian process
surrogate (a fast statistical stand-in for your simulator) to each operating
characteristic, weights each observation by its Monte Carlo noise, scores
candidate designs by expected constrained improvement (how much a candidate
is likely to improve the objective while still satisfying the constraints),
and spends cheap low-replication simulations on exploration while reserving
high-replication runs for verification. It works with any trial simulator,
Bayesian or frequentist, and needs no gradients or closed forms. This
vignette walks through the full workflow: writing an evaluator, running a
calibration, reading its output, sizing the settings, and diagnosing runs
that go wrong.

# Installation

`DiceKriging` and `hetGP` require C and Fortran compilers. On Ubuntu/Debian
install `build-essential` and `gfortran`; on Fedora/RHEL install `gcc`,
`gcc-c++`, and `gcc-gfortran`; on macOS run `xcode-select --install` and add
`gfortran` (from <https://mac.r-project.org/tools/> or `brew install gcc`);
on Windows install Rtools matching your R version. If installation fails
with `gfortran: command not found`, the Fortran compiler is the missing
piece.

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("rashidlab/BATON")

# With vignettes (requires knitr, rmarkdown):
remotes::install_github("rashidlab/BATON", build_vignettes = TRUE)
```

BATON requires R >= 4.2. The companion package
[evolveTrial](https://github.com/naimurashid/evolveTrial) provides
ready-made oncology trial simulators, but it is optional: BATON works with
any evaluator you write yourself, which is what this vignette teaches.

```{r}
library(BATON)
```

# The Evaluator: Your Side of the Contract

BATON's division of labor is strict and simple. You supply one function,
the evaluator, that simulates your trial design at a given set of
parameters and reports its operating characteristics. BATON supplies
everything else: the initial space-filling design, the surrogate models,
the acquisition logic, the fidelity scheduling, and the bookkeeping. To
calibrate your own design you replace exactly one piece of code, the
function that simulates a single trial. Everything in this section is about
getting that one piece right.

## The Contract

The canonical evaluator signature is:

```
function(theta, fidelity, seed = NULL, n_rep = NULL, ...)
```

Each argument has a specific job:

1. **`theta`** is a named list of design parameters (e.g., `theta$nmax`,
   `theta$threshold`). Values arrive exactly as BATON proposes them, which
   means fractional, unless you list a parameter in `integer_params`, in
   which case BATON rounds it before the call. Rounding inside your own
   simulator is the alternative; the two approaches are interchangeable,
   but `integer_params` also makes the returned `best_theta` integer-valued
   for those parameters.
2. **`fidelity`** is a label from `c("low", "med", "high")` naming the
   requested fidelity tier. Multi-fidelity optimization is how BATON
   affords exploration: most evaluations run at the cheap low tier, and the
   expensive high tier is reserved for designs worth verifying.
3. **`n_rep`**: if your function declares an `n_rep` argument, BATON passes
   the exact requested replication count, `fidelity_levels[[fidelity]]`,
   including any dynamic escalation some fidelity methods apply. A
   hardcoded `switch(fidelity, ...)` inside the evaluator is only a
   fallback for calls made outside BATON; if you use one, keep it in sync
   with your `fidelity_levels` or the budget accounting will not match what
   your simulator actually did.
4. **`seed`** is always passed by BATON, derived deterministically from the
   run seed. Apply it with `set.seed(seed)` when non-NULL so every
   evaluation is reproducible. Your own RNG stream is restored when
   `bo_calibrate()` exits.
5. **The return value** is a named numeric vector of operating
   characteristics, for example `c(power = 0.83, type1 = 0.07, EN = 52)`.
   The names are entirely yours; `power`, `type1`, and `EN` are conventions
   of this vignette, not requirements. The only rule is that `objective`
   and every name in `constraints` must be names your evaluator returns.
6. **Variance attributes (recommended)**: attach `attr(result, "variance")`
   (the Monte Carlo variance of each metric's mean) and
   `attr(result, "n_rep")` (the replication count actually used). These
   unlock the noise-aware surrogate discussed below.

## The Minimal Template

The template below is the whole adapter. Part 1 is the only code you write
from scratch; parts 2 and 3 you usually keep as-is.

```{r eval=FALSE}
library(BATON)

# ---- 1. YOUR TRIAL LOGIC (replace this) -------------------------------
# Simulate ONE trial at design parameters theta; return named metrics.
sim_one_trial <- function(theta) {
  # ... your design logic here ...
  stop("REPLACE: simulate one replicate of YOUR design")
  # c(power = ..., type1 = ..., EN = ...)   # any names you like
}

# ---- 2. BATON ADAPTER (usually keep as-is) ----------------------------
my_evaluator <- function(theta, fidelity = c("low", "med", "high"),
                         seed = NULL, n_rep = NULL, ...) {
  fidelity <- match.arg(fidelity)
  if (is.null(n_rep)) {                      # fallback if BATON did not pass n_rep
    n_rep <- switch(fidelity, low = 2000, med = 4000, high = 10000)
  }
  if (!is.null(seed)) set.seed(seed)
  result <- welford_mean_var(function(i, theta) sim_one_trial(theta),
                             n_samples = n_rep, theta = theta)
  metrics <- result$mean
  attr(metrics, "variance") <- result$variance
  attr(metrics, "n_rep") <- result$n
  metrics
}

# ---- 3. CALIBRATE -----------------------------------------------------
fit <- bo_calibrate(
  sim_fun     = my_evaluator,
  bounds      = list(...),        # named list: one c(lower, upper) per parameter
  objective   = "EN",             # a metric your evaluator returns
  constraints = list(power = c("ge", 0.80), type1 = c("le", 0.10)),
  n_init = 20, budget = 60, seed = 1
)
```

Walking through the adapter piece by piece:

- `match.arg(fidelity)` validates the label and picks `"low"` when the
  caller supplies nothing, which is convenient for interactive testing.
- The `switch()` fallback only fires when `n_rep` is `NULL`, i.e., when you
  call the evaluator by hand. When BATON calls it, `n_rep` carries the
  exact count from `fidelity_levels`, so the fallback is bypassed. The
  numbers in the fallback should mirror the `fidelity_levels` you pass to
  `bo_calibrate()` (here the package defaults) so a manual call at
  `fidelity = "low"` behaves like a BATON call at the low tier.
- `welford_mean_var()` runs `sim_one_trial()` `n_rep` times and returns the
  mean and the variance of the mean for every metric in a single pass,
  without ever storing the individual replicates. This matters when
  `n_rep` is large: storing 10,000 replicates of 4 metrics costs about
  312 KB per evaluation, while the Welford accumulator holds two small
  vectors regardless of `n_rep`:

```{r}
n_rep <- 10000; n_metrics <- 4
c(store_all_bytes = 8 * n_rep * n_metrics,
  welford_bytes   = 8 * 2 * n_metrics)
```

- The two `attr()` lines are what turn a plain simulator into a
  noise-aware one. The next section explains exactly what they buy.

## What the Variance and n_rep Attributes Buy You

BATON fits one surrogate per metric, and the surrogate type is chosen per
metric based on what your evaluator provides:

| Scenario | Surrogate used | How to enable |
|----------|----------------|---------------|
| Evaluator attaches `attr(result, "variance")` for the metric | hetGP (recommended) | Use `welford_mean_var()` or compute `var(x)/n_rep` yourself |
| No variance attribute, metric in [0,1] (e.g., power, type I error) | hetGP via binomial approximation | Automatic: BATON estimates `p(1-p)/n_rep` |
| No variance attribute, metric outside [0,1] (e.g., EN) | DiceKriging (homoskedastic) | Provide the variance attribute to upgrade |
| Analytical evaluator (no Monte Carlo noise) | DiceKriging with a small nugget | Appropriate: there is no noise to model |

A heteroskedastic GP (hetGP) lets the model trust precise observations more
than noisy ones. This is not a cosmetic refinement for trial simulators,
because the Monte Carlo variance of expected sample size is strongly
input-dependent: a design with aggressive early stopping produces a bimodal
distribution of realized sample sizes (stopped early vs ran to the end),
hence high variance, while a design that rarely stops early concentrates
near the maximum, hence low variance. A constant-noise GP has to pick one
noise level for the whole space, oversmoothing one region and
undersmoothing the other. In benchmarks, providing the EN variance improved
the best feasible design found by roughly 15%, and variance-aware
surrogates overall reduce the evaluations needed by 30-50%.

If you omit the attributes entirely, nothing breaks. Metrics that live in
[0,1] still get a binomial variance estimate automatically (BATON knows a
proportion's variance from its value and `n_rep`), so power and type I
error surrogates stay noise-aware. Only metrics outside [0,1], typically
your objective, fall back to the homoskedastic model. That is suboptimal
but not incorrect; if you cannot modify an external simulator to report
variances, budget 20-50% more evaluations to compensate.

# A Worked Example: Two-Stage Trial with Futility Stopping

Now we build a complete evaluator from scratch and calibrate it. The design
is a single-arm binary-endpoint trial with one interim look: after a
fraction of the maximum sample size has been enrolled, the trial stops for
futility if the observed response rate is below a futility threshold;
otherwise it enrolls to `n_max` and rejects the null hypothesis if the
final response rate exceeds an efficacy threshold. Four parameters define
the design, and all four are calibrated jointly: `n_max`, the interim
fraction, and the two thresholds.

In practice BATON is aimed at computationally expensive simulators, for
example survival trials with Bayesian posterior-probability stopping rules
that take seconds per evaluation. The simulator here is intentionally fast
so this vignette knits quickly, but the workflow is identical: only the
body of the simulation changes.

## Building the Evaluator

Two modeling decisions come first. The trial tests a null response rate of
0.20 against an alternative of 0.40; these are fixed assumptions, not
calibrated parameters, so they live outside the evaluator. And because
power must be estimated under the alternative while type I error must be
estimated under the null, the evaluator simulates under both hypotheses in
every call and returns both metrics in one vector. This
"both hypotheses per call" pattern is the standard shape for calibration
evaluators: BATON constrains power and type I error simultaneously, so it
needs both numbers for every design it tries.

```{r}
# Fixed assumptions about the trial (not optimized)
p0 <- 0.20  # null response rate
p1 <- 0.40  # alternative response rate

# Two-stage adaptive trial simulator
simple_simulator <- function(theta, fidelity = c("low", "med", "high"),
                             seed = NULL, n_rep = NULL, ...) {
  fidelity <- match.arg(fidelity)

  # BATON passes the exact requested replication count via n_rep. The switch()
  # is only a fallback for calls made outside BATON; keep it in sync with the
  # fidelity_levels used in bo_calibrate() (here the package defaults).
  if (is.null(n_rep)) {
    n_rep <- switch(fidelity,
                    low = 2000,
                    med = 4000,
                    high = 10000)
  }

  if (!is.null(seed)) set.seed(seed)

  # Extract design parameters
  n_max              <- round(theta$n_max)
  interim_fraction   <- theta$interim_fraction
  futility_threshold <- theta$futility_threshold
  efficacy_threshold <- theta$efficacy_threshold

  n_interim <- max(1L, round(interim_fraction * n_max))
  n_final   <- n_max - n_interim

  # --- Vectorised simulation (no loop) ---

  sim_one_hypothesis <- function(p) {
    # Stage 1: interim responses
    x_interim  <- rbinom(n_rep, n_interim, p)
    not_futile <- (x_interim / n_interim) >= futility_threshold

    # Stage 2: remaining responses (only counted for non-futile trials)
    x_remaining <- rbinom(n_rep, n_final, p)
    x_total     <- x_interim + x_remaining

    reject <- not_futile & (x_total / n_max > efficacy_threshold)
    en     <- ifelse(not_futile, n_max, n_interim)

    list(reject = reject, en = en)
  }

  alt  <- sim_one_hypothesis(p1)
  null <- sim_one_hypothesis(p0)

  metrics <- c(
    power = mean(alt$reject),
    type1 = mean(null$reject),
    EN    = mean(alt$en)
  )

  variance <- c(
    power = var(alt$reject)  / n_rep,
    type1 = var(null$reject) / n_rep,
    EN    = var(alt$en)      / n_rep
  )

  attr(metrics, "variance") <- variance
  attr(metrics, "n_rep")    <- n_rep

  return(metrics)
}
```

Design choices worth noting:

- **`round(theta$n_max)`**: sample sizes must be integers but BATON
  proposes continuous values, so the evaluator rounds internally. The
  alternative, `integer_params = "n_max"` in the `bo_calibrate()` call, is
  demonstrated later.
- **Vectorization instead of Welford**: because each replicate here is one
  `rbinom()` draw per stage, all `n_rep` replicates can be simulated in a
  single vectorized call, and the variance of each mean is just
  `var(x) / n_rep`. When one replicate is a genuine function call (a whole
  trial simulation with patient-level logic), use `welford_mean_var()` as
  in the Minimal Template instead. Both routes end at the same place: a
  named metrics vector with `variance` and `n_rep` attributes.
- **`EN` is computed under the alternative** (`mean(alt$en)`). Which
  hypothesis you minimize expected sample size under is a design decision:
  minimizing `E[N]` under the null protects patients when the treatment
  does not work, minimizing it under the alternative shortens trials of
  effective treatments, and either is a legitimate objective. This
  vignette uses the alternative; to switch, return `mean(null$en)`
  instead, or return both as separate metrics and point `objective` at
  either one.

## Sanity-Check the Evaluator Before Optimizing

Always call your evaluator by hand before handing it to `bo_calibrate()`.
This catches shape errors (wrong names, missing attributes) in seconds
instead of twenty minutes into a run, and gives you a feel for the
magnitude of the Monte Carlo noise.

```{r}
test_result <- simple_simulator(
  theta = list(n_max = 60, interim_fraction = 0.5,
               futility_threshold = 0.20, efficacy_threshold = 0.35),
  fidelity = "low",
  seed = 123
)

print(test_result)
print(sqrt(attr(test_result, "variance")))  # standard errors of each metric
```

The output is already informative. This hand-picked design has power of
about 0.73, short of the 0.80 floor, so it is close to but not inside the
feasible region; its type I error is far below 0.10, which says the
efficacy bar of 0.35 is conservative at this sample size; and its expected
sample size is near 60 because futility stopping rarely triggers under the
alternative. In other words, there is room to trade type I error slack for
power, exactly the kind of rebalancing the optimizer will do. The standard
errors confirm the low tier estimates proportions to about a percentage
point here. If your evaluator errors out, returns unnamed values, or
produces `NA`s at this stage, fix that before any optimization.

# Setting Up the Optimization

Three objects define the problem: the parameter box (`bounds`), the metric
to minimize (`objective`), and the conditions a design must satisfy
(`constraints`).

```{r}
# Parameter bounds
bounds <- list(
  n_max              = c(20, 100),   # Maximum sample size
  interim_fraction   = c(0.3, 0.7),  # Fraction enrolled at interim look
  futility_threshold = c(0.05, 0.40),# Stop early if interim rate below this
  efficacy_threshold = c(0.25, 0.60) # Reject H0 if final rate above this
)

# Objective: minimize expected sample size under the alternative
objective <- "EN"

# Constraints: power >= 0.80, type I error <= 0.10
constraints <- list(
  power = c("ge", 0.80),
  type1 = c("le", 0.10)
)
```

**Bounds** should bracket the plausible optimum generously. BATON rescales
everything to the unit cube internally, so wide bounds do not break the
model; they just spend more budget exploring. If the final best design
lands on a bound, that is the signal to widen and re-run. One structural
rule: bounds are independent boxes, so interdependent parameters must be
reparameterized. This example already does it: instead of calibrating an
interim sample size `n_interim` (which must stay below `n_max`, a
constraint a box cannot express), it calibrates `interim_fraction` on
[0.3, 0.7], which is freely variable no matter what `n_max` is. The same
trick applies to ordered thresholds: calibrate the first threshold and a
positive gap rather than two thresholds that must stay ordered.

**Constraints** are a named list mapping a metric name to a direction
(`"ge"` or `"le"`) and a threshold. Leave slack larger than the Monte Carlo
error of your simulator, because a constraint tighter than your noise is
unenforceable. The arithmetic is worth doing explicitly. For a proportion
estimated from `n_rep` replicates, the standard error is
`sqrt(p (1 - p) / n_rep)`:

```{r}
se_type1 <- sqrt(0.10 * 0.90 / 2000)
round(c(se = se_type1, two_se = 2 * se_type1), 4)
```

Estimating a type I error near 0.10 from 2000 replicates gives a standard
error of about 0.0067, so any estimate within roughly +/- 0.013 of the
threshold is statistically indistinguishable from the boundary. Designs in
that band are neither reliably feasible nor reliably infeasible at the low
tier; they need higher-fidelity verification, which is exactly what the
verification step at the end of this vignette provides.

# Running the Calibration

The settings below follow the Quick profile from the Run Profiles table
later in this vignette (`n_init` about 5 x d, `budget` about 15 x d, here
with d = 4 parameters): the right sizing for a first exploratory run while
you iterate on bounds, constraints, and metric definitions. The defaults
(`n_init = 90`, `budget = 300`) are sized for production runs and would be
a lot of simulation to discover a typo in your evaluator.

```{r calibrate, message=FALSE}
fit <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  n_init = 20,      # Initial LHS design points (5 x d, Quick profile)
  q = 2,            # Batch size per iteration
  budget = 60,      # Total evaluations including n_init (15 x d, Quick profile)
  seed = 2025
)
```

## Reading the Progress Output

The chunk above suppresses the live progress messages to keep this
document compact. The transcript below is the actual console output of
exactly that call (same seed, same settings), abridged at the `...` marks.

```
Initialising BATON calibration with seed = 2025
Fidelity selection method: 'adaptive'
  Eval 001 (iter 00, low fidelity): EN = 73.4500
  Eval 002 (iter 00, low fidelity): EN = 96.9820 [feasible]
  Eval 003 (iter 00, low fidelity): EN = 45.6820
  ...
  Eval 019 (iter 00, low fidelity): EN = 45.0000 [feasible]
  Eval 020 (iter 00, low fidelity): EN = 64.9890
Iteration 1: fitting surrogates on 20 evaluations.
  -> max acquisition score: 15.732
  Eval 021 (iter 01, low fidelity): EN = 22.8740
  Eval 022 (iter 01, low fidelity): EN = 27.9760
Iteration 2: fitting surrogates on 22 evaluations.
  -> max acquisition score: 7.183
  Eval 023 (iter 02, low fidelity): EN = 37.5820 [feasible]
  Eval 024 (iter 02, low fidelity): EN = 37.8380 [feasible]
  -> best observed feasible: 37.582 (change: -7.418)
  ...
Iteration 7: fitting surrogates on 32 evaluations.
  -> max acquisition score: 6.421
  Eval 033 (iter 07, low fidelity): EN = 27.7795 [feasible]
  Eval 034 (iter 07, low fidelity): EN = 32.4280 [feasible]
  -> best observed feasible: 27.779 (change: -9.803)
  ...
Iteration 16: fitting surrogates on 50 evaluations.
  -> max acquisition score: 0.930
  Eval 051 (iter 16, low fidelity): EN = 26.6400 [feasible]
  Eval 052 (iter 16, low fidelity): EN = 24.9230
  -> best observed feasible: 26.640 (change: -0.297)
  -> No improvement > 0.1% in last 5 iterations (count: 1)
  ...
Iteration 20: fitting surrogates on 58 evaluations.
  -> max acquisition score: 0.199
  Eval 059 (iter 20, low fidelity): EN = 25.1840
  Eval 060 (iter 20, low fidelity): EN = 23.4960
  -> best observed feasible: 26.640 (change: 0.000)
```

Reading it line by line:

- **`Initialising ... seed = 2025`** confirms the run seed. Per-evaluation
  seeds are derived from it, so the whole run is reproducible.
- **`Eval 001 (iter 00, low fidelity): EN = 73.4500`**: iteration 00 is
  the initial Latin hypercube design, 20 space-filling points evaluated
  before any modeling happens. Each line reports the objective value; the
  `[feasible]` tag appears when the evaluation satisfied every constraint.
  Only 4 of the 20 initial points are feasible here, and the feasible ones
  have large `EN` (the easy way to get power is a big trial). That is the
  typical starting picture: feasibility is easy to buy with sample size,
  and the optimization's job is to find the cheapest design that still
  qualifies.
- **`Iteration 1: fitting surrogates on 20 evaluations`**: the BO loop
  proper begins. Each iteration refits the GP surrogates on everything
  seen so far, scores a candidate pool with the acquisition function, and
  evaluates the `q = 2` best candidates.
- **`-> max acquisition score: 15.732`**: the acquisition value of the
  most promising candidate. Watch this number across iterations: it
  starts large (much apparent room for improvement) and decays toward
  zero, with some wobble, as the surrogates firm up and the incumbent
  gets hard to beat. A sustained collapse to nearly zero is one of
  BATON's stopping signals (`status = "acq_flatline"`).
- **`-> best observed feasible: 37.582 (change: -7.418)`**: the incumbent,
  the best feasible objective so far, improved from 45.0 (found in the
  initial design) to 37.58. The big drops at iterations 2 and 7 and the
  small one at 16 are the classic BO trajectory: large early gains, then
  refinement, then a plateau. From iteration 16 onward the incumbent does
  not move; the run spends its remaining budget confirming there is
  nothing better nearby, and ends when the 60-evaluation budget is
  consumed.
- **`-> No improvement > 0.1% in last 5 iterations (count: 1)`**: the
  early-stopping counter. Two consecutive stagnant patience windows end
  the run with `status = "early_stopped"`; here the counter was reset
  once (the iteration-16 improvement was still inside the lookback
  window) and the budget ran out before a second stagnant window
  accumulated.
- **`low fidelity` on every line**: the adaptive fidelity method judged
  the cheap tier's information sufficient throughout this small, fast
  problem. On harder problems, or with `fidelity_costs` reflecting a truly
  expensive high tier, you will see `med` and `high` evaluations appear as
  the run turns from exploration to verification. Either way, the pattern
  is why the final high-fidelity verification step later in this vignette
  is part of the workflow rather than an optional extra.

## The Result Object

```{r}
fit$best_theta
fit$best_objective
fit$status
```

`best_theta` is the best feasible design found: a trial with a maximum of
about 28 patients, an interim look at about 42% of enrollment, a futility
bar of about 0.23, and an efficacy bar of about 0.29. `best_objective` is
its expected sample size under the alternative. Note the fractional
`n_max`: BATON optimized a continuous parameter and the evaluator rounded
internally, so before acting on this design you round it and re-check, as
shown in the verification section. `status` records why the run ended;
`"budget_exhausted"` means the full 60 evaluations were spent, the normal
ending for a budget-sized run.

Feasibility here is operational and deterministic: a design is marked
`feasible` when its *observed* metrics satisfy every constraint, in this
case observed power >= 0.80 and observed type I error <= 0.10 in that
evaluation's own replicates. It is a statement about Monte Carlo
estimates, not about the truth; a design whose true power is 0.79 can be
marked feasible by a lucky batch of replicates, and vice versa. This is
why constraint slack should exceed the Monte Carlo error and why
verification at high fidelity matters for designs near a boundary.

There is no `best_metrics` field; the metrics of the best design are read
from the history, which stores one unpacked column per metric:

```{r}
feasible <- fit$history[fit$history$feasible, ]
best_row <- feasible[which.min(feasible$objective), ]
best_row[, c("power", "type1", "EN")]
```

The best design's observed power (0.818) is within about two standard
errors of the 0.80 boundary at the low tier (`sqrt(0.8 * 0.2 / 2000)` is
about 0.009), exactly the near-boundary situation the constraints section
warned about. The verification section below settles it.

## The History

Every evaluation the run performed is one row of `fit$history`:

```{r}
names(fit$history)
```

The columns fall into four groups: bookkeeping (`iter`, `eval_id`,
`theta_id`, `fidelity`, `n_rep`), the design itself (`theta` and `unit_x`
as list-columns, plus one unpacked column per parameter), the results
(`metrics` and `variance` as list-columns, one unpacked column per metric,
and `objective`), and the optimizer's view of the point (`feasible`,
`prob_feas` the surrogate's feasibility probability when the point was
proposed, `cv_estimate` its coefficient of variation, `acq_score` its
acquisition value). The unpacked columns make ordinary data manipulation
easy:

```{r}
head(fit$history[, c("iter", "fidelity", "n_rep", "objective", "feasible")])
table(fit$history$fidelity)
sum(fit$history$n_rep)  # total simulated trial replicates (per hypothesis)
```

Two diagnostics worth glancing at after any run: the fraction of feasible
evaluations (here `r sum(fit$history$feasible)` of
`r nrow(fit$history)`; very low fractions suggest constraints near the
edge of what the design space supports) and where the feasible points sit
in objective terms.

## Two Pictures of the Run

With `ggplot2` installed, two built-in plots summarize the geometry of the
run. The feasible frontier plot shows every evaluation in the
(EN, power) plane, colored by feasibility. You can see the trade-off
directly: power rises with expected sample size, the feasible points are
those above the power floor, and the optimization has pushed into the
bottom-left corner of the feasible set, small trials that still clear 80%
power.

```{r feasible-frontier, fig.alt="Scatter plot of power against expected sample size, colored by feasibility"}
plot_feasible_frontier(fit)
```

The trade-off surface plot shows the same evaluations in the
(type I error, power) plane, colored by objective value. The darker
(low-EN) points cluster where both error rates are acceptable but not
wastefully small, illustrating why minimizing EN pushes designs toward,
but not across, the constraint boundaries.

```{r tradeoff, fig.alt="Scatter plot of power against type I error, colored by objective value"}
plot_tradeoff_surfaces(fit)
```

# Choosing Your Settings

This table is the sizing reference for every knob, with `d` the number of
design parameters. The paragraphs after it explain the failure modes each
recommendation avoids.

| Setting | Default | How to choose |
|---------|---------|---------------|
| `n_init` | 90 | Space-filling evaluations before BO starts. 10 x `d` for a working run; 4-5 x `d` for a quick pass. |
| `budget` | 300 | Total evaluations including `n_init`. 20-30 x `d` for well-behaved problems; add 50-100% when constraints are tight. |
| `q` | 2 | Evaluations proposed per BO iteration. 2-4; larger batches give wall-clock parallelism at a small cost in statistical efficiency. |
| `bounds` | required | Bracket the plausible optimum generously; widen and re-run if the best design lands on a bound. Reparameterize interdependent parameters. |
| `constraints` | required | Leave slack larger than the Monte Carlo error; verify near-boundary designs at higher fidelity. |
| `fidelity_levels` | `c(low = 2000, med = 4000, high = 10000)` | Size the low tier cheap enough for exploration and the high tier precise enough for verification. |
| `integer_params` | `NULL` | Names of parameters BATON rounds before calling your simulator. |
| `seed` | 2025 | Run seed; per-evaluation seeds derive from it. |

**`n_init`** buys the surrogates their first, unbiased look at the design
space. Too small and two things go wrong: the GP fit is poorly conditioned
(you may see surrogate fitting warnings, or lengthscales pinned at their
bounds), and the first BO iterations chase noise in a model that has no
real information yet, spending budget on points an adequate initial design
would have ruled out. Too large and you waste budget that adaptive
selection would have spent better; 10 x `d` is the standard compromise,
and the default of 90 corresponds to a d = 9 problem at that rate. On an
expensive simulator, remember that omitting `n_init` means 90 initial
evaluations before the first BO iteration happens at all.

**`budget`** is the total evaluation count, including the initial design,
so `budget = 60` with `n_init = 20` leaves 40 BO evaluations, 20
iterations at `q = 2`. The 20-30 x `d` guidance assumes the feasible
region is a reasonable fraction of the box. When constraints carve out
only a sliver, the optimizer first has to find the region before it can
optimize within it, which is what the 50-100% surcharge pays for. The
history tells you which situation you are in: if the first feasible point
appears late and feasible rows stay rare, the budget is mostly funding the
search for feasibility.

**`q`** trades statistical efficiency for wall-clock time. With `q > 1`,
BATON selects a spatially diverse batch (local penalization) rather than
`q` near-duplicates of the top candidate, and with `options(BATON.cores)`
set on Unix/macOS the batch is evaluated in parallel. Each point after the
first in a batch is chosen with slightly staler information, which is the
small efficiency cost.

**`fidelity_levels`** should be sized from the constraint arithmetic. The
low tier absorbs most evaluations, so it must be cheap; but it also
decides which designs look feasible during the search, so its standard
error should not dwarf your constraint slack. The high tier is the
verification instrument: size it so the standard error is comfortably
below the slack you care about. Four selection methods are available via
`fidelity_method`:

| Method | One-line description |
|--------|----------------------|
| `"adaptive"` (default) | Cost-aware: maximizes expected information gain per unit cost |
| `"staged"` | Iteration-based schedule: low early, CV/boundary-triggered promotion mid-run, high late |
| `"threshold"` | Legacy: P(feasible) >= 0.75 high, >= 0.40 med, else low |
| `"hybrid_staged"` | MCEM-inspired staging with CV-based escalation and dynamic high-tier scaling |

The default adaptive method computes, for each proposed design, a value
score (acquisition value, objective uncertainty, and proximity to the
feasibility boundary) and divides by the tier's cost raised to an exponent
that grows as the run progresses and the budget depletes; the tier with
the best value per unit cost wins. When computational cost does not scale
linearly with replications (I/O overhead, parallel startup), pass
`fidelity_costs` to tell the selector the true relative prices. Custom
tiers are just a named vector; because `simple_simulator` declares an
`n_rep` argument, BATON passes these exact counts to it and the internal
`switch()` fallback is bypassed:

```{r custom-fidelity, message=FALSE, warning=FALSE}
fit_custom <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  fidelity_levels = c(low = 500, med = 2000, high = 5000),
  n_init = 12,
  budget = 30,
  seed = 11
)
table(fit_custom$history$fidelity)
```

As in the main run, the adaptive selector kept this cheap problem entirely
at the low tier; the chunk demonstrates the mechanics, and on an expensive
simulator the same call produces a mixed tier allocation.

**`integer_params`** moves the rounding from your evaluator into BATON.
The practical difference shows up in the answer:

```{r integer-params, message=FALSE}
fit_int <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  n_init = 20, q = 2, budget = 60, seed = 2025,
  integer_params = "n_max"
)
fit_int$best_theta$n_max
fit_int$status
```

`best_theta$n_max` is now an integer, directly usable as a protocol
number. This run also happens to demonstrate early stopping: the
incumbent stopped improving and the run ended as `"early_stopped"` (no
relative improvement above 0.1% over a 5-iteration patience window, twice
in a row; configure or disable via the `early_stop` argument) rather than
spending its full budget. Rounding coarsens the search space, so distinct
proposals can map to the same evaluated design, which tends to bring
convergence forward.

**`seed`** makes runs reproducible end to end. BATON derives a
deterministic per-evaluation seed from the run seed and passes it to your
evaluator, and restores your session's RNG stream on exit, so a
calibration does not perturb other random work in your script.

## Run Profiles: Quick, Balanced, Thorough

Rather than reasoning about each knob from scratch every time, start from
one of three profiles:

| Setting | Quick | Balanced | Thorough |
|---------|-------|----------|----------|
| `n_init` | 4-5 x d | 10 x d | 10 x d |
| `budget` | 10-15 x d | 20-30 x d | 40-50 x d or more |
| `fidelity_levels` | `c(low = 200, med = 500, high = 2000)` | default | raise high tier (e.g., `high = 50000`) |
| `q` | 2-4 | 2 | 2 |
| `early_stop` | default | default | `list(enabled = FALSE)` or tightened `threshold` |
| `multi_seed_verify` | `FALSE` | `TRUE` | `TRUE` with `multi_seed_n = 10` |

Use Quick while you iterate on the problem formulation: debugging the
evaluator, taking a first look at a new design space, checking that
feasibility is achievable at all, adjusting bounds and constraints. Runs
take minutes, and the low tier's precision is coarse; at 200 replicates a
proportion near 0.8 has standard error `sqrt(0.8 * 0.2 / 200)`, about
0.028, so operating characteristics are good to roughly +/- 2-3 percentage
points. Switch to Balanced for the run whose answer you intend to act on:
overnight or lunch-scale depending on your simulator, with high-tier
standard errors around 0.003-0.005 at the default 10,000 replicates.
Reserve Thorough for manuscript or regulatory numbers, for constraints
sitting within about 2 standard errors of a boundary, and for d >= 6; at
`high = 50000` the standard error drops to about 0.002:

```{r}
round(c(quick_low   = sqrt(0.8 * 0.2 / 200),
        balanced_high = sqrt(0.8 * 0.2 / 10000),
        thorough_high = sqrt(0.8 * 0.2 / 50000)), 4)
```

Move between profiles as the problem firms up; the worked example above is
a Quick-profile run, which is why its answer gets a verification pass
before anyone acts on it.

# When Things Go Wrong

## Nothing Is Feasible

The clearest failure mode is a constraint set the design space cannot
satisfy. Here is what it looks like: the same problem as above, but
demanding 99% power with 1% type I error.

```{r infeasible-demo, warning=TRUE, message=FALSE}
fit_bad <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = list(power = c("ge", 0.99), type1 = c("le", 0.01)),
  n_init = 12, q = 2, budget = 24, seed = 7,
  progress = FALSE
)
any(fit_bad$history$feasible)
max(fit_bad$history$power)
min(fit_bad$history$type1)
```

Three things to notice. First, the warning (emitted at each internal
best-point lookup, so it can print more than once): when no evaluation
satisfies all constraints, BATON warns and returns the best *infeasible*
point in `best_theta`, so the fit object is still usable as a starting
point rather than an error. Second, the diagnosis is in the history:
designs in this space get close to 99% power (0.989 observed), and
separately reach essentially zero type I error, but no evaluated design
achieved both at once, so the joint requirement is what fails. Third, `prob_feas` in the history stays low across the run,
the surrogate's own way of saying the feasible region looks empty.

What to do: check the constraint thresholds are achievable at all (a
larger `n_max` upper bound would help here), expand the bounds in the
direction the near-feasible points are piling up, or relax the thresholds.
If you believe the region exists but is small, increase the budget; the
expected-violation logic steers the search toward the feasibility boundary
even before the first feasible point is found. When in doubt, run the
infeasible fit's best point through your simulator at high `n_rep` and
look at how far each constraint actually is from its threshold.

## The Best Design Sits on a Bound

If any component of `best_theta` is at (or hard against) the edge of its
box, the optimum probably lies outside the box, and the reported best is
just where the search ran out of room. Widen that bound and re-run,
optionally passing the previous run's evaluations via
`initial_history = fit$history` so the earlier budget is not wasted. The
`refine_bounds()` helper automates the opposite move, shrinking bounds
around the best region for a second-stage run; see
`vignette("advanced-features")`.

## Reading fit$status

`fit$status` always records why the run ended:

| `status` | Meaning |
|----------|---------|
| `budget_exhausted` | Normal completion: the evaluation budget was consumed |
| `early_stopped` | Converged: no meaningful improvement over the patience window |
| `acq_flatline` | Acquisition values collapsed; no candidate looks promising |
| `walltime` | The `max_walltime_s` cap tripped; result is whatever was found in time |
| `cancelled` | Your `callback` requested cancellation |
| `errored` | A failure was caught under `on_error = "return_partial"`; see `fit$error_message` |

`budget_exhausted` and `early_stopped` are the two healthy endings.
`acq_flatline` after very few iterations suggests the surrogates see no
structure, worth checking that your metrics actually vary across the
bounds. `walltime`, `cancelled`, and `errored` all return partial but
internally consistent fits: the history contains every completed
evaluation, and `best_theta` is the best of what was seen.

## Surrogate Fitting Failures

An error like `Failed to fit surrogate for metric ...` means the GP's
covariance matrix was ill-conditioned, usually because the design points
are too few or too clustered, or because a variance estimate was zero or
non-finite. Increase `n_init` for a more diverse initial design, and check
your evaluator's variance attribute is positive and finite for every
metric (a metric that is constant across replicates, for example type I
error estimated as exactly 0 from a small `n_rep`, produces a zero
variance; more replicates or the binomial fallback both avoid this).

## Performance Seems Slow

First check whether the time is going into your simulator or into BATON's
modeling; on all but toy problems the simulator dominates, and the fix is
parallelism inside the evaluator (see Scaling Up below). If the modeling
side is slow, note the candidate pool is sized adaptively at 500 x `d`
points, clamped to [1000, 5000], with a 50% increase in the final 30% of
iterations; the `candidate_pool` argument only sets a floor (default
2000), so lowering it below the adaptive size has no effect.

# Before You Trust the Answer

Three checks before a calibrated design goes anywhere important.

## 1. Verify at High Fidelity Across Seeds

Most evaluations during a run happen at low fidelity, and the incumbent
was selected partly because its noise broke in its favor. Setting
`multi_seed_verify = TRUE` adds a verification stage after the
optimization: the best design is re-evaluated at `multi_seed_n`
independent seeds (default 5) at high fidelity
(`multi_seed_reps`, default `fidelity_levels[["high"]]`), and the fit
gains a verdict. The call below is the worked example's calibration with
only that flag added; the optimization itself is identical because the
seed is the same.

```{r verify, message=FALSE}
fit_verified <- bo_calibrate(
  sim_fun = simple_simulator,
  bounds = bounds,
  objective = objective,
  constraints = constraints,
  n_init = 20, q = 2, budget = 60, seed = 2025,
  multi_seed_verify = TRUE
)
fit_verified$verdict
# multi_seed_summary is a named list; pull out the cross-seed means and sds
round(unlist(fit_verified$multi_seed_summary[
  c("n_seeds", "n_rep", "power_mean", "power_sd",
    "type1_mean", "type1_sd", "EN_mean", "EN_sd")]), 4)
```

`MULTI_SEED_PASS` means the design was feasible at every verification
seed: at 10,000 replicates per seed, power holds at about 0.814 (above
0.80) and type I error at about 0.080 (below 0.10), with cross-seed
standard deviations small enough to rule out a fluke. The other verdicts
are `MULTI_SEED_FAIL` (at least one seed violated a constraint, under the
default `multi_seed_strict = TRUE`) and `MULTI_SEED_WARN` (the same
failure with `multi_seed_strict = FALSE`). A failed verdict on a
near-boundary design is common and informative: it usually means the
optimizer exploited low-fidelity noise, and the remedy is a slightly
slacker constraint, a larger budget, or a Thorough-profile re-run.

## 2. Round Integer Parameters and Re-Check

Unless you passed `integer_params`, `best_theta` comes back fractional.
Round to the design you would actually run, then re-simulate at large
`n_rep` to confirm the rounded design still meets the constraints:

```{r round-recheck}
theta_final <- fit$best_theta
theta_final$n_max <- round(theta_final$n_max)

final_check <- simple_simulator(theta_final, seed = 999, n_rep = 100000)
round(final_check, 4)
```

At 100,000 replicates the standard errors are around 0.001, so this is as
close to ground truth as simulation gets: the rounded design keeps power
above 0.80 and type I error below 0.10. If rounding had pushed a
constraint over its threshold (it can, when the optimum sits near a
boundary), the fix is to check the neighboring integer designs by hand or
re-run with `integer_params` so the optimizer only ever visits integer
designs.

## 3. Check fit$status

Confirm the run ended for a healthy reason (`"budget_exhausted"` or
`"early_stopped"`). A `"walltime"` or `"cancelled"` result is whatever was
found in the time available, not a converged answer, and an `"errored"`
fit should be inspected via `fit$error_message` before any of its numbers
are reused.

# Scaling Up

## Parallelism Inside the Evaluator

The main opportunity for parallelism is inside your evaluator: splitting
Monte Carlo replicates across cores so each evaluation completes faster.
`pool_welford_results()` combines chunk-level Welford accumulators with
Chan's algorithm, preserving exact means and variances:

```{r eval=FALSE}
parallel_simulator <- function(theta, fidelity = "high", n_cores = 4,
                               n_rep = NULL, ...) {
  # Fallback for calls outside BATON; keep in sync with your fidelity_levels
  if (is.null(n_rep)) {
    n_rep <- switch(fidelity, low = 2000, med = 4000, high = 10000)
  }
  chunk_size <- ceiling(n_rep / n_cores)

  # NOTE: mclapply() is POSIX-only (no forking on Windows). Windows users
  # should use a PSOCK cluster (parallel::makeCluster() + parLapply()).
  chunks <- parallel::mclapply(1:n_cores, function(core_id) {
    welford_mean_var(
      sample_fn = function(i, theta) {
        # REPLACE with one replicate of YOUR trial; return a named vector
        stop("REPLACE: simulate one replicate of your design")
      },
      n_samples = chunk_size,
      theta = theta
    )
  }, mc.cores = n_cores)

  pooled <- pool_welford_results(chunks)

  metrics <- pooled$mean
  attr(metrics, "variance") <- pooled$variance
  attr(metrics, "n_rep") <- pooled$n
  metrics
}
```

Each chunk gets `ceiling(n_rep / n_cores)` replicates, so the total may
slightly exceed `n_rep`; that is harmless, and reporting `pooled$n` keeps
the accounting honest. Separately, `options(BATON.cores = k)` evaluates
the points *within* a batch in parallel on Unix/macOS, which pays off when
`q > 1`; the two levels of parallelism multiply, so budget your cores
across them. `vignette("variance-estimation")` covers the Welford
machinery, its numerical properties, and an honest benchmark of its
overhead.

## Service and Long-Run Controls

For unattended, scheduled, or cluster runs, `bo_calibrate()` has controls
for capping walltime, checkpointing, surviving simulator failures, and
cooperative cancellation; details for each are in `?bo_calibrate`.

```{r eval=FALSE}
fit <- bo_calibrate(
  sim_fun = simple_simulator, bounds = bounds, objective = "EN",
  constraints = constraints, n_init = 20, budget = 60, seed = 42,
  max_walltime_s = 6 * 3600,    # graceful stop before the scheduler kills the job
  checkpoint_fun = function(snap) saveRDS(snap, "checkpoint.rds"),
  checkpoint_every = 5,         # snapshot every 5 BO iterations
  on_error = "return_partial",  # keep completed evaluations if the simulator fails
  callback = function(info) !file.exists("STOP"),  # return TRUE to continue
  slim = TRUE                   # lightweight return: skips the final surrogate refit
)

# Resume a stopped run from its checkpoint
ck <- readRDS("checkpoint.rds")
fit <- bo_calibrate(
  sim_fun = simple_simulator, bounds = bounds, objective = "EN",
  constraints = constraints, initial_history = ck$history,
  n_init = nrow(ck$history), budget = nrow(ck$history) + 20, seed = 43
)
```

Every one of these endings produces a valid `BATON_fit` whose `status`
field says what happened, so downstream code can branch on it.

# Where to Go Next

This vignette covered the core workflow: one evaluator, one calibration,
verified. The package's other vignettes each go deeper on one axis:

- **`vignette("BATON-methods")`**: how the optimizer works internally, GP
  surrogates and their noise models, the expected constrained improvement
  acquisition, batch selection by local penalization, and the
  multi-fidelity selection methods. Read this to understand *why* the
  settings above are sized the way they are.
- **`vignette("BATON-case-study")`**: an end-to-end realistic calibration,
  including sensitivity analysis (Sobol indices, gradients), benchmarking
  BO against grid and random search, and calibrating a family of designs
  under different design philosophies.
- **`vignette("advanced-features")`**: multi-stage workflows,
  `refine_bounds()`, warm-starting from saved state, and fidelity control.
- **`vignette("variance-estimation")`**: Welford's algorithm, pooling with
  `pool_welford_results()`, and when the variance attributes matter most.
- **`vignette("v04-hardening")`**: cross-philosophy warm-starts and the
  multi-seed verification gate in depth.

For analyses beyond a single calibration, the package also ships
`benchmark_methods()` (compare BO against grid/random/heuristic
strategies), `estimate_constraint_reliability()` (large-sample validation
of constraint satisfaction across repeated calibrations),
`ablation_multifidelity()` (compare fidelity policies), and the
`sa_sobol()` / `sa_gradients()` / `summarise_case_study()` family for
sensitivity analysis and reporting; the case-study vignette demonstrates
them in context. See also the project README on GitHub for release notes
and installation troubleshooting, and
`system.file("examples/simulator_with_variance.R", package = "BATON")` for
additional evaluator patterns.

# Session Info

```{r}
sessionInfo()
```
