---
title: "v0.4.0 hardening: cross-philosophy warmstart and multi-seed verification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{v0.4.0 hardening: cross-philosophy warmstart and multi-seed verification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Background

BATON v0.4.0 introduces two safeguards to `bo_calibrate()`:

1. **`warmstart_from`**: a Stage 0 parameter that injects previously
   calibrated designs as initial seeds alongside the Latin hypercube
   sample, helping the GP surrogate escape local minima.
2. **`multi_seed_verify`**: a Stage 4 verification gate that re-evaluates
   the calibrated design at multiple independent seeds at high fidelity,
   detecting stored-vs-actual operating-characteristic gaps.

Plus a new batch-level wrapper `bo_calibrate_philosophies()` that
orchestrates both safeguards across a panel of philosophies.

## Why these were added

Both features address empirical failure modes documented in the
bounds-widening campaign (May 2026) for the JASA 2026 paper on adaptive
trial calibration:

1. **BO local minima.** A multi-arm Optimal calibration with $w_0 = 1$
   (sole objective: minimize $E_0[N]$) terminated at $E_0[N] = 67.19$,
   while a Fleming calibration with $w_0 = 0.5$ terminated at
   $E_0[N] = 66.18$. Fleming Pareto-dominated Optimal under Optimal's own
   loss — a calibration artifact attributable to BO settling at a local
   minimum in the 9-parameter design space. Seeding Stage 0 with
   Fleming's top-3 high-fidelity verified designs recovered Optimal at
   $E_0[N] = 65.66$, strictly better than both prior calibrations.

2. **Stored-vs-actual gaps.** Three SYMM cohortA bidirectional designs
   had calibration-time stored type-I error below the 0.10 regulatory
   floor but multi-seed verification (5 seeds × R = 10,000 each)
   revealed multi-seed type-I error in [0.122, 0.170] — stored-vs-actual
   gaps of approximately 0.075, more than 20 Monte Carlo standard
   errors. Without multi-seed verification, those designs would have
   shipped type-I-inflated.

Both failure modes are caught by the new safeguards.

## Stage taxonomy

| Stage | Purpose | Fidelity (R) | Seed |
|-------|---------|--------------|------|
| Stage 1 | BO exploration | LOW (e.g., 3000) | single |
| Stage 2 | BO refinement | MED (e.g., 5000) | single |
| Stage 3 | High-fidelity verification at the calibrated design | HIGH (e.g., 10000) | single |
| Stage 4 (v0.4.0) | Multi-seed verification at the calibrated design | HIGH | multiple |

Stage 1/2/3 are **fidelity tiers** within a single BO run at a single
seed. They harden the surrogate's belief about the calibrated design
against within-run uncertainty.

Stage 4 (v0.4.0) is on an **orthogonal axis**: seed-dependent
variability of the calibrated design's true operating characteristics.
It hardens the calibration against between-trajectory uncertainty.

## Single-philosophy usage

```{r}
library(BATON)

fit <- bo_calibrate(
  sim_fun = your_simulator,
  bounds = list(
    total_n = c(20, 200),
    eff = c(0.85, 0.99),
    fut = c(0.05, 0.40),
    beat = c(1.0, 16),
    ev = c(3, 10),
    margin = c(0.0, 1.0),
    prior_strength = c(0.1, 1.0),
    n_intervals = c(2, 5)
  ),
  objective = "weighted_EN_N",
  constraints = list(power = c("ge", 0.80),
                     type1 = c("le", 0.10)),
  budget = 300,
  fidelity_levels = c(low = 3000, med = 5000, high = 10000),
  multi_seed_verify = TRUE,
  multi_seed_n = 5,
  multi_seed_strict = TRUE
)

# Inspect Stage 4 results
fit$verdict                            # "MULTI_SEED_PASS" or "MULTI_SEED_FAIL"
fit$multi_seed_summary$strict_feas     # 1 if all seeds pass, 0 otherwise
fit$multi_seed_summary$type1_mean      # seed-averaged type-I
fit$multi_seed_runs                    # per-seed evaluations

# Gate adoption on Stage 4 verdict
if (fit$verdict == "MULTI_SEED_PASS") {
  adopt(fit$best_theta)
} else {
  message("Stage 4 verification failed; do not adopt.")
  print(fit$multi_seed_summary)
}
```

## Cross-philosophy warmstart

When a philosophy's calibration appears stuck in a local minimum, seed
its Stage 0 with a sibling philosophy's calibrated design:

```{r}
# Calibrate Fleming first (no warmstart needed).
fit_fleming <- bo_calibrate(
  sim_fun = your_simulator,
  bounds = your_bounds,
  objective = "weighted_EN_N",
  constraints = list(power = c("ge", 0.80), type1 = c("le", 0.10)),
  budget = 300,
  multi_seed_verify = TRUE
)

# Calibrate Optimal with Fleming's design as a Stage 0 seed.
fit_optimal <- bo_calibrate(
  sim_fun = your_simulator,
  bounds = your_bounds,
  objective = "EN_null",                 # Optimal's loss
  constraints = list(power = c("ge", 0.80), type1 = c("le", 0.10)),
  warmstart_from = list(fit_fleming$best_theta),
  budget = 300,
  multi_seed_verify = TRUE
)

# Optimal should now have access to Fleming's basin via the surrogate.
```

The `warmstart_from` argument also accepts paths to `*_best_design.csv`
files produced by prior calibrations:

```{r}
fit_optimal <- bo_calibrate(
  sim_fun = your_simulator,
  bounds = your_bounds,
  objective = "EN_null",
  constraints = list(power = c("ge", 0.80), type1 = c("le", 0.10)),
  warmstart_from = list("results/pipeline/fleming_run/fleming_run_best_design.csv"),
  budget = 300,
  multi_seed_verify = TRUE
)
```

## Multi-philosophy batch calibration

For Pareto-frontier comparison papers, use the new batch wrapper:

```{r}
batch <- bo_calibrate_philosophies(
  scenario_id = "cohortA_strong_effect",
  sim_fun = your_simulator,
  bounds = your_bounds,
  philosophies = list(
    Fleming = list(objective = "weighted_EN_N",
                    constraints = list(power = c("ge", 0.80),
                                       type1 = c("le", 0.10))),
    Minimax = list(objective = "total_n",
                    constraints = list(power = c("ge", 0.80),
                                       type1 = c("le", 0.10))),
    Admissible = list(objective = "weighted_EN_N",
                       constraints = list(power = c("ge", 0.80),
                                          type1 = c("le", 0.10))),
    Optimal = list(objective = "EN_null",
                    constraints = list(power = c("ge", 0.80),
                                       type1 = c("le", 0.10))),
    `Alt-Optimal` = list(objective = "EN_alt",
                          constraints = list(power = c("ge", 0.80),
                                             type1 = c("le", 0.10)))
  ),
  budget = 300,
  multi_seed_verify = TRUE,
  multi_seed_strict = TRUE
)

# The manifest documents which philosophies passed multi-seed and what
# they were warmstarted from.
batch$manifest
#>     philosophy dependency_role          verdict n_max objective_value warmstart_donors  runtime_s
#> 1      Fleming           donor MULTI_SEED_PASS    51         13.520                       1234.5
#> 2      Minimax           donor MULTI_SEED_PASS    47         15.213                        987.6
#> 3   Admissible           donor MULTI_SEED_PASS    49         13.881                       1156.3
#> 4      Optimal       recipient MULTI_SEED_PASS    63         11.380   Fleming;Minimax;Admissible  1289.7
#> 5  Alt-Optimal       recipient MULTI_SEED_PASS    51         16.260   Fleming;Minimax;Admissible  1342.1
```

## Backward compatibility

All v0.4.0 changes are additive and opt-in. Existing calls to
`bo_calibrate()` continue to work unchanged:

- `multi_seed_verify` defaults to `FALSE` (Stage 4 not run).
- `warmstart_from` defaults to `NULL` (LHS-only Stage 0).
- The pre-existing `initial_history` parameter still works; `warmstart_from`
  is a higher-level convenience layer that converts user-friendly design
  specifications to `initial_history` internally. Specifying both in the
  same call is an error.

## Recommended workflow for production / regulatory calibrations

1. **Always set `multi_seed_verify = TRUE` and `multi_seed_strict = TRUE`.**
   The performance cost is minor (~20% of total wall time at typical
   budgets); the failure-detection benefit is substantial.
2. **For multi-philosophy work, prefer `bo_calibrate_philosophies()`** over
   ad hoc orchestration of `bo_calibrate()` calls. The wrapper handles
   dependency ordering, warmstart provenance, and the manifest of
   verdicts.
3. **Gate adoption on `fit$verdict == "MULTI_SEED_PASS"`** rather than
   only on Stage 3 verification. The stored-vs-actual gap can be large
   (we observed 0.075 on type-I) and is rule-variant- and
   basin-specific.
4. **If a calibration fails Stage 4, recalibrate** rather than adopting
   a flagged design. Possible recalibration strategies (in order of
   effort):
   - Adjust bounds (smaller search box may help BO converge to a more
     stable basin).
   - Increase budget (more BO iterations may find a stabler design).
   - Warmstart from a sibling philosophy via `warmstart_from`.
   - Switch random seed and rerun.
5. **Document which BO seed produced which design.** BO under wide
   bounds with bounded budget exhibits restart variance — running the
   same calibration with a different seed can produce a different
   calibrated design. Multi-seed verification certifies the seeds the
   calibration was tested at, not all possible seeds.
